\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{xurl}
\usepackage{enumitem}
\usepackage{microtype}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!50!black,
  urlcolor=blue!55!black,
  citecolor=blue!50!black
}

\input{build_info.tex}

\newcommand{\R}{\mathbb{R}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\diag}{\operatorname{diag}}

\title{Prediction-Synchronized Local Polynomial Smoothing}
\author{%
  \small
  \begin{tabular}{@{}c@{}}
  Source path: \texttt{\string~/current\_projects/geosmooth/split\_handoffs/}\\
  \texttt{lps\_prediction\_synchronized\_design\_2026-06-05/}\\
  \texttt{lps\_prediction\_synchronized\_design.tex}
  \end{tabular}%
}
\date{Build: \reportbuilddatetime}

\begin{document}
\maketitle

\section*{Purpose}

This design note defines the first version of prediction-synchronized local
polynomial smoothing, abbreviated \emph{PS-LPS}.  The hyphenated name is used
throughout this note because it remains readable in prose and parallels names
such as S-LPL-TF.  The goal is to keep the empirical strength and simplicity
of local polynomial smoothing (LPS), while reducing the incoherence that can
occur when many overlapping local polynomial charts are fitted independently.

The starting observation is that ordinary LPS constructs one local model around
each anchor point.  These models overlap heavily: many charts make predictions
on the same data points.  Ordinary LPS uses this overlap only indirectly.  The
proposed synchronized version uses it directly by penalizing disagreement
between local predictions on shared overlap points.

The first implementation should be deliberately modest:
\begin{itemize}[leftmargin=*]
  \item no trend-filtering \(\ell_1\) penalty;
  \item no adaptive local parameter field yet;
  \item fixed candidate parameters \((k,\kappa,p,d)\) inside each fit;
  \item one quadratic synchronization parameter
    \(\lambda_{\mathrm{sync}}\ge 0\);
  \item sparse least-squares implementation whenever possible.
\end{itemize}
This makes PS-LPS a natural bridge between ordinary LPS
and S-LPL-TF / SLPLiFT.  It is easier than full SLPLiFT because the objective is
purely quadratic, but it tests the same core idea: local models should agree
where their prediction domains overlap.

\section{Ordinary LPS Notation}

Let
\[
  x_1,\ldots,x_n \in \R^D
\]
be the observed covariate points, and let
\[
  y_i = f(x_i) + \varepsilon_i
\]
be the observed response at point \(i\).  The ambient dimension is \(D\), but
the local chart dimension may be smaller.

For each anchor \(i\), ordinary LPS chooses a support
\[
  N_i \subset \{1,\ldots,n\},
  \qquad |N_i| = k,
\]
usually the \(k\) nearest neighbors of \(x_i\).  A kernel \(\kappa\) assigns
nonnegative weights
\[
  w_{ij} \ge 0,
  \qquad j\in N_i.
\]
The weight \(w_{ij}\) is large when \(x_j\) is close to anchor \(x_i\), and is
small when \(x_j\) lies near the edge of the local support.

At anchor \(i\), construct local coordinates
\[
  z_{ij}\in\R^d,
  \qquad j\in N_i,
\]
for example by local PCA.  Let
\[
  \phi_{i}(x_j) = \phi(z_{ij})\in\R^q
\]
be the local polynomial feature vector of degree \(p\).  For degree two,
\(\phi(z)\) contains the intercept, linear terms, squares, and cross terms.
The number of columns is
\[
  q = 1+d+\frac{d(d+1)}{2}.
\]

Ordinary LPS fits a separate coefficient vector
\[
  \beta_i\in\R^q
\]
at each anchor by weighted least squares:
\[
  \widehat\beta_i^{\mathrm{LPS}}
  =
  \argmin_{\beta_i}
  \frac12
  \sum_{j\in N_i}
  w_{ij}\{y_j-\phi_i(x_j)^\top\beta_i\}^2.
\]
The fitted value at the anchor is usually
\[
  \widehat f_i^{\mathrm{LPS}}
  =
  \phi_i(x_i)^\top\widehat\beta_i^{\mathrm{LPS}}.
\]
Because each \(\widehat\beta_i^{\mathrm{LPS}}\) is fitted independently,
neighboring charts may disagree even when they make predictions on the same
overlap points.

\section{Prediction Synchronization}

PS-LPS keeps one local coefficient vector
\[
  \beta_i\in\R^q
\]
for each anchor, but estimates all of them jointly.  The local prediction made
by chart \(i\) at point \(r\in N_i\) is
\[
  g_i(x_r) = \phi_i(x_r)^\top\beta_i.
\]
If another chart \(\ell\) also contains \(x_r\), then chart \(\ell\) makes its
own prediction
\[
  g_\ell(x_r) = \phi_\ell(x_r)^\top\beta_\ell.
\]
The synchronization idea is to penalize
\[
  g_i(x_r)-g_\ell(x_r)
\]
on shared overlap points.

Define the overlap between charts \(i\) and \(\ell\) as
\[
  O_{i\ell}=N_i\cap N_\ell.
\]
Only pairs with \(O_{i\ell}\ne\varnothing\) can contribute synchronization
rows.  The basic quadratic synchronization penalty is
\[
  \mathcal S(\beta)
  =
  \frac12
  \sum_{(i,\ell)\in\mathcal E_{\mathrm{sync}}}
  \sum_{r\in O_{i\ell}}
  \omega_{i\ell r}
  \left\{
    \phi_i(x_r)^\top\beta_i
    -
    \phi_\ell(x_r)^\top\beta_\ell
  \right\}^2.
\]
Here \(\mathcal E_{\mathrm{sync}}\) is the set of chart pairs that are allowed
to synchronize, and \(\omega_{i\ell r}\ge 0\) is the overlap weight assigned to
the shared prediction point \(r\).

\section{Recommended Overlap Weights}

The natural first definition uses the product of the two chart weights:
\[
  \omega_{i\ell r}^{\mathrm{raw}}
  =
  w_{ir}w_{\ell r},
  \qquad r\in O_{i\ell}.
\]
This is attractive for exactly the reason discussed in the planning thread.  A
point \(x_r\) near the center of both charts receives a large product weight.
A point near the boundary of one or both charts receives a small product
weight.  Thus the penalty asks charts to agree most strongly in the middle of
their shared predictive region, and it is less strict near the edge of either
local chart.

The raw product is conceptually clean, but its scale changes with the number of
overlap points and with the kernel.  For the first implementation, the
recommended default is the pair-normalized product weight
\[
  \omega_{i\ell r}
  =
  |O_{i\ell}|
  \frac{
    w_{ir}w_{\ell r}
  }{
    \sum_{s\in O_{i\ell}} w_{is}w_{\ell s}+\epsilon
  },
  \qquad r\in O_{i\ell}.
\]
The small constant \(\epsilon>0\) prevents division by zero.  With this
normalization,
\[
  \sum_{r\in O_{i\ell}}\omega_{i\ell r}
  \approx |O_{i\ell}|.
\]
So the average overlap weight within a chart pair is approximately one, while
the product structure still emphasizes the middle of the overlap.

This choice separates two roles:
\begin{itemize}[leftmargin=*]
  \item the relative weights inside one overlap are determined by the product
    \(w_{ir}w_{\ell r}\);
  \item the global strength of synchronization is controlled by
    \(\lambda_{\mathrm{sync}}\), not by accidental kernel scale.
\end{itemize}
For auditability, the implementation should allow the raw product as an option,
but the first experiments should use the normalized product as the default.

\section{Objective Function}

For fixed candidate parameters
\[
  \theta=(k,\kappa,p,d)
\]
and synchronization strength \(\lambda_{\mathrm{sync}}\), define
\[
  \widehat\beta(\theta,\lambda_{\mathrm{sync}})
  =
  \argmin_{\beta_1,\ldots,\beta_n}
  \left[
    \mathcal L_{\mathrm{local}}(\beta)
    +
    \lambda_{\mathrm{sync}}\mathcal S(\beta)
  \right],
\]
where
\[
  \mathcal L_{\mathrm{local}}(\beta)
  =
  \frac12
  \sum_{i=1}^{n}
  \sum_{j\in N_i}
  w_{ij}
  \{y_j-\phi_i(x_j)^\top\beta_i\}^2.
\]
The local loss says that each chart should fit the observed responses in its
own support.  The synchronization term says that overlapping charts should
produce similar predictions on shared points.

When
\[
  \lambda_{\mathrm{sync}}=0,
\]
the mathematical synchronization term is inactive.  If no additional
stabilization is used, the problem decomposes into ordinary independent LPS
chart fits.  If an explicit ridge stabilization is used, the
\(\lambda_{\mathrm{sync}}=0\) problem instead decomposes into independent
ridge-stabilized LPS chart fits.  The implementation and reports must state
which case is being used.  As \(\lambda_{\mathrm{sync}}\) increases, the local
models are increasingly forced to agree on overlaps.  This is the key
difference from ordinary LPS.

\section{Ridge-Stabilized Implementation Contract}

The first implementation now treats numerical ridge stabilization as an
explicit modeling and reporting parameter rather than as an invisible solver
detail.  For a nonnegative ridge scale
\(\lambda_{\mathrm{ridge}}\), the implemented objective is
\[
  \mathcal L_{\mathrm{local}}(\beta)
  +
  \lambda_{\mathrm{sync}}\mathcal S(\beta)
  +
  \frac12
  \sum_{i=1}^{n}
  \rho_i
  \|\beta_i\|_2^2,
\]
where \(\rho_i\) is a scale-relative ridge used for chart \(i\).  In the
prototype, \(\rho_i\) is derived from the local weighted-design scale and the
reported user parameter \(\lambda_{\mathrm{ridge}}\).

This gives two separate nesting checks:
\begin{itemize}[leftmargin=*]
  \item if \(\lambda_{\mathrm{ridge}}=0\) and
    \(\lambda_{\mathrm{sync}}=0\), PS-LPS should reproduce ordinary
    independent LPS fixed-chart fitted values;
  \item if \(\lambda_{\mathrm{ridge}}>0\) and
    \(\lambda_{\mathrm{sync}}=0\), PS-LPS should reproduce independent
    ridge-stabilized LPS fixed-chart fitted values.
\end{itemize}
Therefore ridge-stabilized PS-LPS must be compared against a matched
ridge-stabilized LPS baseline before interpreting any remaining improvement as
a synchronization effect.

\section{Total Local GCV Inside PS-LPS}

The earlier independent-LPS experiment showed that selecting global parameters
by minimizing
\[
  \sum_{i=1}^{n}\operatorname{GCV}_i
\]
for independently fitted local charts is not a reliable parameter-selection
rule on the first-batch examples.  PS-LPS should nevertheless record the same
kind of diagnostic, because the object being measured is now different.

After solving the synchronized problem, chart \(i\) has coefficients
\(\widehat\beta_i\).  Its synchronized local support residual sum of squares is
\[
  \operatorname{RSS}^{\mathrm{PS}}_i
  =
  \sum_{j\in N_i}
  w_{ij}
  \{y_j-\phi_i(x_j)^\top\widehat\beta_i\}^2.
\]
Let \(r_i\) be a local complexity proxy, initially the numerical rank of the
weighted local design matrix for chart \(i\).  Define the synchronized local
GCV diagnostic as
\[
  \operatorname{GCV}^{\mathrm{PS}}_i
  =
  \frac{\operatorname{RSS}^{\mathrm{PS}}_i/k}
  {(1-r_i/k)^2}.
\]
The total synchronized local GCV is
\[
  \operatorname{GCV}^{\mathrm{PS}}_{\Sigma}
  =
  \sum_{i=1}^{n}
  \operatorname{GCV}^{\mathrm{PS}}_i.
\]
This is \emph{not} the same as the total local GCV of ordinary LPS.  The
residuals are computed from prediction-synchronized charts, so they include the
effect of overlap agreement.  For this reason, the diagnostic may be more
connected with cross-validation error than the independent-LPS version, even
though it should still be treated as a diagnostic rather than as the primary
selection rule in the first implementation.

\section{Sparse Least-Squares Form}

The objective is quadratic.  Therefore the first implementation can be framed
as one sparse least-squares problem.

Stack all local coefficients into
\[
  b =
  (\beta_1^\top,\ldots,\beta_n^\top)^\top \in \R^{nq}.
\]
Every local data-fit row has the form
\[
  \sqrt{w_{ij}}\,
  \phi_i(x_j)^\top\beta_i
  \approx
  \sqrt{w_{ij}}\,y_j.
\]
This row touches only the block of variables belonging to chart \(i\).

Every synchronization row has the form
\[
  \sqrt{\lambda_{\mathrm{sync}}\omega_{i\ell r}}\,
  \left\{
    \phi_i(x_r)^\top\beta_i
    -
    \phi_\ell(x_r)^\top\beta_\ell
  \right\}
  \approx 0.
\]
This row touches only two blocks of variables: chart \(i\) and chart \(\ell\).
Thus the global design matrix is sparse even though the number of local charts
is \(n\).

\section{Prediction Rule}

The primary fitted value should preserve ordinary LPS semantics:
\[
  \widehat f_i
  =
  \phi_i(x_i)^\top\widehat\beta_i.
\]
That is, the prediction at \(x_i\) is read from the chart anchored at \(i\).
This makes direct comparison with ordinary LPS clean.

The implementation may also record an overlap-averaged diagnostic prediction
\[
  \widehat f_i^{\mathrm{avg}}
  =
  \frac{
    \sum_{\ell:\,i\in N_\ell} v_{\ell i}
    \phi_\ell(x_i)^\top\widehat\beta_\ell
  }{
    \sum_{\ell:\,i\in N_\ell} v_{\ell i}
  },
\]
where \(v_{\ell i}\) can be the chart weight \(w_{\ell i}\).  This is useful
diagnostically, but it should not be the primary fitted value in the first
experiment because it changes two things at once: synchronization and prediction
aggregation.

\section{Cross-Validation Without Response Leakage}

For cross-validation, validation responses must not enter the local data-fit
loss.  If \(V\) is a validation fold, then data-fit rows with response index
\[
  j\in V
\]
should receive zero response weight in
\(\mathcal L_{\mathrm{local}}\).  The synchronization rows may still include
validation covariate locations because they use only \(X\), chart bases, and
kernel weights.  This does not leak validation responses.

The held-out prediction at \(x_j\), \(j\in V\), is computed after solving the
fold-specific synchronized least-squares problem.  The CV score is then the
usual held-out RMSE over validation responses.  The candidate grid should
include
\[
  \lambda_{\mathrm{sync}}\in
  \{0,\lambda_1,\lambda_2,\ldots\},
\]
so the ordinary LPS case is always present as a nested reference.

\section{First Implementation Defaults}

The first implementation should be intentionally narrow.

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Object & First default & Reason \\
\midrule
Local chart & ordinary local PCA & matches current LPS path \\
Polynomial degree & degree \(2\) & current preferred LPS setting \\
Chart dimension & fixed \(d\), plus later auto/local.auto & simplest solver contract \\
Overlap pairs & nearby anchor pairs with nonempty overlap & avoids all-pairs blowup \\
Overlap weight & normalized product \(w_{ir}w_{\ell r}\) & emphasizes overlap interior \\
Prediction & anchor prediction \(\phi_i(x_i)^\top\beta_i\) & comparable to LPS \\
Solver & sparse QR/least-squares path & objective is quadratic \\
\bottomrule
\end{tabular}
\end{center}

The first validation should use the frozen first-batch non-manifold datasets
where ordinary LPS \texttt{auto} and \texttt{local.auto} results already exist.
This keeps the first comparison focused:
\[
  \text{ordinary LPS}
  \quad\text{versus}\quad
  \text{prediction-synchronized LPS}.
\]

\section{Diagnostics}

The first experiment should report:
\begin{itemize}[leftmargin=*]
  \item selected \(\lambda_{\mathrm{sync}}\), support size, kernel, degree, and
    chart dimension;
  \item Truth RMSE, Observed RMSE, and CV RMSE where synthetic truth is
    available;
  \item total synchronized local GCV
    \[
      \operatorname{GCV}^{\mathrm{PS}}_{\Sigma}
      =
      \sum_i \operatorname{GCV}^{\mathrm{PS}}_i;
    \]
  \item synchronization energy
    \[
      \mathcal S(\widehat\beta);
    \]
  \item local prediction disagreement before and after synchronization;
  \item whether pointwise Truth-RMSE contribution maps improve in regions where
    \texttt{local.auto} previously lost to \texttt{auto};
  \item runtime and sparse-system size diagnostics.
\end{itemize}

The most important scientific question is not whether synchronization always
improves the average score on the first attempt.  The first question is whether
the method changes the right object: does it reduce incoherent overlap
predictions without collapsing into oversmoothing or interpolation?

\section{Out of Scope for the First Version}

Several extensions are attractive but should not be mixed into the first
implementation:
\begin{itemize}[leftmargin=*]
  \item an \(\ell_1\) trend-filtering penalty;
  \item locally varying support size, kernel, degree, or chart dimension;
  \item GCV-driven adaptive local parameter fields;
  \item overlap-averaged prediction as the main estimator;
  \item second-order local SVD charts;
  \item theoretical rate claims.
\end{itemize}
These can be added later if the quadratic synchronized LPS core behaves
sensibly.

\section{Summary}

Prediction-synchronized LPS estimates all local polynomial charts jointly.  The
new ingredient is the quadratic penalty
\[
  \sum_{(i,\ell)}
  \sum_{r\in O_{i\ell}}
  \omega_{i\ell r}
  \left\{
    \phi_i(x_r)^\top\beta_i
    -
    \phi_\ell(x_r)^\top\beta_\ell
  \right\}^2.
\]
The recommended first overlap weight is the normalized product of the two
ordinary chart weights.  This keeps the middle-of-overlap intuition while
making the scale of the synchronization parameter easier to compare across
datasets and candidate settings.

The immediate next step is an implementation phase that builds the sparse
quadratic system for fixed
\((k,\kappa,p,d,\lambda_{\mathrm{sync}},\lambda_{\mathrm{ridge}})\), verifies
the two lambda-zero nesting checks above, and then runs a small grid over
\(\lambda_{\mathrm{sync}}\) on the frozen first-batch datasets with ordinary
LPS and ridge-matched LPS baselines kept separate.

\end{document}
