---
title: "Finding your way around geosmooth"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Finding your way around geosmooth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1120px; }
td, th { vertical-align: top; }
td code { white-space: normal; overflow-wrap: anywhere; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`geosmooth` estimates a response or probability mass on coordinate data and
graphs. Start with the task you want to perform, then choose a method. You do
not need to learn the operator constructors or synthetic-data helpers to fit
your first model.

This guide describes the current API of version 0.2.0. The catalog below covers
all 60 explicitly exported functions, each defined in `geosmooth`. The 25
geometry and sampling helpers formerly re-exported here are now called directly
from `dgraphs`; a separate dependency catalog below lists them.
The package also registers 44 S3 methods, which R selects when you call a generic
such as `predict()` or `print()` on an appropriate object. Those are described
by workflow rather than as 44 additional entry points.

## Choose a starting point

| Your task | Start here | What to know before choosing |
|---|---|---|
| Regress a response on coordinates | [`fit.lps()`](https://pgajer.github.io/geosmooth/reference/fit.lps.html) | Local polynomial smoothing; selects neighborhood size, degree, and kernel by cross-validation. |
| Average predictions from overlapping local models | [`fit.malps()`](https://pgajer.github.io/geosmooth/reference/fit.malps.html) | Model-averaged local polynomial smoothing; supports coordinate and graph-geodesic neighborhoods. |
| Couple local polynomial predictions across overlaps | [`fit.ps.lps()`](https://pgajer.github.io/geosmooth/reference/fit.ps.lps.html) | Prediction-synchronized local polynomial smoothing; experimental R reference implementation. |
| Fit a locally adaptive trend with a sparse difference penalty | [`fit.lpl.tf()`](https://pgajer.github.io/geosmooth/reference/fit.lpl.tf.html) | Local polynomial lifting trend filtering; requires optional package `genlasso`. |
| Add synchronization to a lifting trend filter | [`fit.slpl.tf()`](https://pgajer.github.io/geosmooth/reference/fit.slpl.tf.html) | Synchronized local polynomial lifting trend filtering; adds a second, quadratic penalty. |
| Smooth a response already attached to graph vertices | [`fit.metric.graph.lowpass()`](https://pgajer.github.io/geosmooth/reference/fit.metric.graph.lowpass.html) or [`fit.graph.trend.filtering()`](https://pgajer.github.io/geosmooth/reference/fit.graph.trend.filtering.html) | Choose spectral low-pass smoothing or a graph-difference penalty. Edge-weight meanings differ; see below. |
| Penalize local second derivatives on a point cloud | [`fit.ssrhe.hessian.regression()`](https://pgajer.github.io/geosmooth/reference/fit.ssrhe.hessian.regression.html) | Hessian-energy regression, following semi-supervised regression using Hessian energy (SSRHE); use `lambda.selection = "cv"` or `"gcv"` to select penalties. |
| Use an absolute-value Hessian penalty | [`fit.ssrhe.hessian.l1.regression()`](https://pgajer.github.io/geosmooth/reference/fit.ssrhe.hessian.l1.regression.html) | Different regularization from quadratic Hessian energy; has its own solver and tuning controls. |
| Interpolate graph values while fixing a region boundary | [`perform.harmonic.smoothing()`](https://pgajer.github.io/geosmooth/reference/perform.harmonic.smoothing.html) | Iterative harmonic relaxation; [`harmonic.smoother()`](https://pgajer.github.io/geosmooth/reference/harmonic.smoother.html) also records extrema and topology diagnostics. |
| Estimate probability mass over a fixed support | [`fit.density()`](https://pgajer.github.io/geosmooth/reference/fit.density.html) | Takes nonnegative counts or masses; returns mass over the rows of `X`. |
| Estimate where one subject visits that support | [`fit.subject.od()`](https://pgajer.github.io/geosmooth/reference/fit.subject.od.html) | Takes visit indices, including repeats; handles occupation-density fitting and optional visit-level validation. |
| Build reproducible synthetic examples | [`synthetic.spec()`](https://pgajer.github.io/geosmooth/reference/synthetic.spec.html) then [`materialize.synthetic()`](https://pgajer.github.io/geosmooth/reference/materialize.synthetic.html) | Combine geometry, sampling, truth, and response noise. |

These choices describe different estimators, not a ranking of predictive
performance. Parallel-transport trend filtering and transported graph Hessians
have their own [experimental section](#experimental-transport-methods).

## A first regression fit

Use `X` with one observation per row and a response `y` of matching length.
A local chart is a coordinate system fitted near an observation, often using
principal component analysis (PCA). Start in the supplied coordinates before
adding local charts or changing numerical controls.

```{r first-fit}
library(geosmooth)
set.seed(1)
X <- matrix(seq(0, 1, length.out = 60), ncol = 1)
y <- sin(2 * pi * X[, 1]) + rnorm(nrow(X), sd = 0.08)
fit <- fit.lps(
  X, y,
  foldid = rep(1:5, length.out = nrow(X)),
  support.grid = c(8L, 12L, 16L),
  degree.grid = 0:1,
  kernel.grid = "gaussian"
)
head(predict(fit))
new.X <- matrix(seq(0.1, 0.9, length.out = 5), ncol = 1)
predict(fit, newdata = new.X)
```

Cross-validation (CV) holds out responses to choose tuning parameters. A
support is the neighborhood used for a local fit; a grid lists candidate
values. The local support size and polynomial degree jointly determine whether
a local fit has enough observations. For grouped observations, use
[`lps.grouped.foldid()`](https://pgajer.github.io/geosmooth/reference/lps.grouped.foldid.html) to keep each group in one fold. Use [`lps.nested.cv()`](https://pgajer.github.io/geosmooth/reference/lps.nested.cv.html) when
you need an outer assessment of the whole tuning procedure.

### Prediction and reusable fits

| Fitted object | How to get predictions | Scope |
|---|---|---|
| Local polynomial smoother | `predict(fit)` or `predict(fit, newdata = Xnew)` | Existing evaluation points or new coordinate rows. |
| Model-averaged local polynomial smoother | `predict(fit)` or `predict(fit, newdata = Xnew)` | New coordinates are supported for coordinate-distance fits; graph-geodesic fits currently reject new targets. |
| Either lifting trend filter | `predict(fit)` | Training points only; `newdata` must be `NULL`. |
| Prediction-synchronized, chart-kernel, local-likelihood, graph low-pass, graph trend-filtering, Hessian, or parallel-transport fit | `fit$fitted.values` | No registered `predict()` method for these classes in this version. Chart-kernel and local-likelihood fits can receive `X.eval` when fitting. |
| Density or occupation-density result | `fit$rho` | Probability mass on the evaluation support. |
| Harmonic smoothing result | `fit$harmonic_predictions` | Values on the original graph vertices. |

A *refit* changes the response while reusing geometry or an operator. It does
not automatically rerun the original model-selection procedure. Each refit
method documents whether penalties remain fixed or can be selected again.
An *operator* describes the differences or derivatives being penalized; it is
not itself a fitted response. A *spectral basis* stores graph eigenvectors so
many responses can be smoothed without recomputing them.

### Refit and extract smoother matrices

Use `refit(fit, y = new.y, ...)` for all six reusable model families.
R chooses the method from the fitted object's class. The controls in the
last column belong to that method; they are not shared by every model.

| Fitted model | Accepted response | Geometry reused and tuning behavior |
|---|---|---|
| MALPS | Vector; omitted or `NULL` reuses the stored response | Supports and averaging weights stay fixed. Optional `weights` affect local coefficient estimation; robust residual weights are recomputed. |
| Lifting trend filter | Vector required | Reuses the residual operator and stored penalty, or accepts a fixed `lambda`. |
| Synchronized lifting trend filter | Vector required | Reuses both operators and stored penalties, or accepts fixed `lambda1` and `lambda2`. |
| Metric graph low-pass | Vector or matrix required | Reuses the spectral basis and smoothing parameter; `per.column.gcv = TRUE` selects a parameter for each response column. |
| Quadratic Hessian-energy regression | Vector or matrix; omitted or `NULL` reuses the stored response | Reuses the operator and fixed penalties; accepts `lambda1`, `lambda2`, `weights`, and `ridge`. CV/GCV fits dispatch here and retain their selected penalties by default. |
| Hessian L1 regression | One response; omitted or `NULL` reuses the stored response | Reuses the derivative operator; accepts `lambda.grid`, fixed or CV `lambda.selection`, and solver controls. |

For repeated metric graph low-pass or quadratic Hessian refits, keep using
the original fitted object. Their returned refit summaries cannot themselves
be refitted. The other four families return reusable fitted objects.

Use `smoother.matrix(fit, ...)` for LPS and MALPS. LPS accepts `check.tol`
and returns a map from training responses to evaluation predictions under
its fixed-configuration restrictions. MALPS accepts `max.n` and
`allow.robust` and returns a training-point matrix conditional on stored
supports and weights. For robust MALPS, `allow.robust = TRUE` freezes the final
robust weights; it does not capture their response-dependent changes.

```{r shared-generics}
malps.fit <- fit.malps(
  X, y, degree = 1L, support.type = "knn", support.size = 12L
)
new.y <- cos(2 * pi * X[, 1])
malps.refit <- refit(malps.fit, y = new.y)
S <- smoother.matrix(malps.fit)
max(abs(as.vector(S %*% new.y) - predict(malps.refit)))
```

The former six `refit.*()` entry points and two `*.smoother.matrix()` entry
points are no longer exported. Replace `fitted.model =` with `object =` and
`y.new =` with `y =` in old refit calls. Other method-specific arguments and
return classes are unchanged. `help("refit", package = "geosmooth")` and
`help("smoother.matrix", package = "geosmooth")` link to each method's full
argument list; former help names remain available.

## Function catalog

Each row in the geosmooth catalog gives one exported name and its purpose. In an R session,
open the corresponding help with, for example,
`help("fit.lps", package = "geosmooth")`. The final two sections explicitly
list related functions owned by `dgraphs`; they are not geosmooth exports, and
their detailed help lives in dgraphs.

### Shared model operations

| Function | Purpose |
|---|---|
| [`refit()`](https://pgajer.github.io/geosmooth/reference/refit.html) | Refit a supported model with a new response using stored geometry and method-specific tuning controls. |
| [`smoother.matrix()`](https://pgajer.github.io/geosmooth/reference/smoother.matrix.html) | Extract the conditional linear map for a supported LPS or MALPS fit. |

### Local regression and prediction synchronization

| Function | Purpose |
|---|---|
| [`fit.lps()`](https://pgajer.github.io/geosmooth/reference/fit.lps.html) | Fit and tune local polynomial regression, including supported Gaussian and binary response modes. |
| [`fit.malps()`](https://pgajer.github.io/geosmooth/reference/fit.malps.html) | Fit and optionally tune a model-averaged collection of local polynomial models. |
| [`fit.ps.lps()`](https://pgajer.github.io/geosmooth/reference/fit.ps.lps.html) | Fit experimental local polynomial models whose predictions are synchronized on overlaps. |
| [`fit.chart.kernel()`](https://pgajer.github.io/geosmooth/reference/fit.chart.kernel.html) | Fit a chart-kernel field, with optional reference-measure quadrature weights. |
| [`fit.local.likelihood()`](https://pgajer.github.io/geosmooth/reference/fit.local.likelihood.html) | Fit local density or Bernoulli likelihood models; returns a fitted field. |

### Model assessment and uncertainty

| Function | Purpose |
|---|---|
| [`lps.grouped.foldid()`](https://pgajer.github.io/geosmooth/reference/lps.grouped.foldid.html) | Assign whole groups of observations to validation folds. |
| [`lps.nested.cv()`](https://pgajer.github.io/geosmooth/reference/lps.nested.cv.html) | Assess a tuned local polynomial smoother using outer and inner validation folds. |
| [`lps.backend.diagnostics()`](https://pgajer.github.io/geosmooth/reference/lps.backend.diagnostics.html) | Inspect the computational backend, chart dimensions, and selected controls. |
| [`lps.pointwise.band()`](https://pgajer.github.io/geosmooth/reference/lps.pointwise.band.html) | Calculate pointwise uncertainty for a supported fixed-configuration Gaussian fit. |
| [`malps.gcv()`](https://pgajer.github.io/geosmooth/reference/malps.gcv.html) | Calculate generalized cross-validation and optional leave-one-out diagnostics for a stored model-averaged fit. |
| [`bootstrap.malps()`](https://pgajer.github.io/geosmooth/reference/bootstrap.malps.html) | Measure refit variability by bootstrap case weights while holding the selected geometry fixed. |

Generalized cross-validation (GCV) uses a smoother's effective degrees of
freedom to estimate prediction error without explicitly fitting every
leave-one-out model. It is not interchangeable with every form of CV.

The two uncertainty workflows have different assumptions. LPS uncertainty
requires a Gaussian fit, the R backend, the documented orthogonal polynomial
basis, a single candidate configuration, and a fixed chart dimension. Its
pointwise band also requires evaluation at the training points. MALPS matrix
and bootstrap diagnostics condition on the selected supports and weights;
they do not repeat model selection. Neither workflow provides simultaneous
coverage or automatic uncertainty for a full data-selected pipeline.

### Lifting trend filters

| Function | Purpose |
|---|---|
| [`lpl.tf.operator()`](https://pgajer.github.io/geosmooth/reference/lpl.tf.operator.html) | Build a local polynomial residual operator. |
| [`fit.lpl.tf()`](https://pgajer.github.io/geosmooth/reference/fit.lpl.tf.html) | Fit a trend filter with an absolute-value penalty on those residuals. |
| [`slpl.tf.operator()`](https://pgajer.github.io/geosmooth/reference/slpl.tf.operator.html) | Build the residual operator together with an overlap synchronization operator. |
| [`fit.slpl.tf()`](https://pgajer.github.io/geosmooth/reference/fit.slpl.tf.html) | Fit the residual penalty plus quadratic synchronization. |

Both fitters accept a prebuilt operator or can construct one. For a single fit,
start with the fitter. Build operators explicitly when comparing responses,
studying neighborhoods, or reusing expensive geometry. Both fitters currently
require `genlasso`; their supported prediction domain is the training set.

### Graph spectral smoothing and graph trend filtering

| Function | Purpose |
|---|---|
| [`fit.metric.graph.lowpass()`](https://pgajer.github.io/geosmooth/reference/fit.metric.graph.lowpass.html) | Select a graph-spectral smoothing parameter by GCV and fit a response. |
| [`metric.graph.lowpass.operator()`](https://pgajer.github.io/geosmooth/reference/metric.graph.lowpass.operator.html) | Convert metric edge lengths to conductances and construct a graph Laplacian. |
| [`metric.graph.lowpass.basis()`](https://pgajer.github.io/geosmooth/reference/metric.graph.lowpass.basis.html) | Construct and retain graph eigenvectors for repeated smoothing. |
| [`apply.metric.graph.lowpass.path()`](https://pgajer.github.io/geosmooth/reference/apply.metric.graph.lowpass.path.html) | Apply a supplied basis over a parameter grid; retain the candidate fits. |
| [`metric.graph.heat.eta.grid()`](https://pgajer.github.io/geosmooth/reference/metric.graph.heat.eta.grid.html) | Choose heat-time candidates using spectral information. |
| [`metric.graph.heat.extend.lower()`](https://pgajer.github.io/geosmooth/reference/metric.graph.heat.extend.lower.html) | Propose a smaller heat time for an external search, subject to resolution safeguards. |
| [`graph.trend.filtering.operator()`](https://pgajer.github.io/geosmooth/reference/graph.trend.filtering.operator.html) | Construct graph differences or experimental path divided differences. |
| [`fit.graph.trend.filtering()`](https://pgajer.github.io/geosmooth/reference/fit.graph.trend.filtering.html) | Fit an absolute-value graph-difference penalty; requires `genlasso`. |

Graph construction and shortest paths are supplied by `dgraphs`. A graph's
adjacency list contains 1-based vertex indices; a parallel weight list gives
the corresponding edge values. **Check their meaning for the chosen method:**
metric low-pass smoothing interprets them as lengths and converts them to
conductances. Recursive graph trend filtering uses conductance-like weights,
whereas its path divided-difference option uses lengths. Conductance measures
connection strength; it is not the same quantity as distance.

Heat time (`eta`) controls the amount of spectral smoothing. A truncated basis
can fail to resolve small positive heat times; use the returned spectral
diagnostics when constructing or extending a path yourself.

### Hessian-energy regression

| Function | Purpose |
|---|---|
| [`ssrhe.hessian.operator()`](https://pgajer.github.io/geosmooth/reference/ssrhe.hessian.operator.html) | Estimate local derivative operators and Hessian-energy matrices from point-cloud neighborhoods. |
| [`fit.ssrhe.hessian.regression()`](https://pgajer.github.io/geosmooth/reference/fit.ssrhe.hessian.regression.html) | Fit quadratic Hessian-energy regression using fixed penalties, label CV, or GCV; optionally select supports. |
| [`fit.ssrhe.hessian.l1.regression()`](https://pgajer.github.io/geosmooth/reference/fit.ssrhe.hessian.l1.regression.html) | Fit an absolute-value derivative penalty with fixed or CV-selected regularization. |
| [`ssrhe.support.grid()`](https://pgajer.github.io/geosmooth/reference/ssrhe.support.grid.html) | Build candidate adaptive-radius support profiles for tuning. |

The Hessian describes local second derivatives. A quadratic penalty spreads
curvature reduction across the fitted field; an absolute-value penalty can
set derivative components to zero. These are different estimators.

Label CV supports semi-supervised use: some vertices may lack an observed
response. GCV currently requires one fully observed response vector and
strictly positive observation weights. Exact GCV traces and randomized trace
estimates have different computational costs and reproducibility properties.
For the absolute-value fitter, `solver = "admm"` selects the implemented
alternating direction method of multipliers; the `genlasso` solver requires
that optional package.

To select quadratic Hessian penalties, use the same fitter with an explicit
selection mode. Mode-specific settings belong in `cv.control` or `gcv.control`.
Scalar penalties and candidate grids cannot be combined.

```{r hessian-selection}
hessian.fit <- fit.ssrhe.hessian.regression(
  X, y, k = 12L, tangent.dim = 1L,
  lambda.selection = "cv", lambda1.grid = c(0.01, 0.1),
  cv.control = list(foldid = rep(1:5, length.out = nrow(X)))
)
hessian.fit$lambda
```

The former `.cv()` and `.gcv()` fitter names have been retired. Their old help
names still lead to the unified fitter's migration instructions. Existing
fixed calls with `lambda1` and optional `lambda2` continue to work.

### Harmonic interpolation

| Function | Purpose |
|---|---|
| [`perform.harmonic.smoothing()`](https://pgajer.github.io/geosmooth/reference/perform.harmonic.smoothing.html) | Relax interior values while holding the region boundary fixed; return convergence diagnostics. |
| [`harmonic.smoother()`](https://pgajer.github.io/geosmooth/reference/harmonic.smoother.html) | Run relaxation while recording extrema, prediction history, and topology stability. |
| [`get.region.boundary()`](https://pgajer.github.io/geosmooth/reference/get.region.boundary.html) | Find region vertices with a neighbor outside the region. |

Use `print()`, `summary()`, and `plot()` on the tracked `harmonic_smoother`
object. The two smoothing functions currently differ in boundary-free behavior
so they should not be treated as aliases. In the
current native implementation, boundary vertices are defined by a neighbor
outside the region; a degree-one vertex inside a whole-graph region is not
automatically fixed. The basic function returns unchanged input with
`status = "no_boundary"` when no boundary exists; the tracked function still
runs relaxation. Both require small value changes and a small harmonic residual
for numerical convergence. Extrema stability is a separate diagnostic:
`stability_detected` records whether a window was found, and `stable_iteration`
is `NA` otherwise. On an alternating path, recording every second iteration
can appear stable while the values continue to oscillate. Check `converged`
and `max_residual`; a stability detection does not establish optimal smoothing.

### Probability mass and subject occupation

| Function | Purpose |
|---|---|
| [`fit.density()`](https://pgajer.github.io/geosmooth/reference/fit.density.html) | Fit empirical or graph-random-walk mass estimates from nonnegative weights on a support. |
| [`normalize.density()`](https://pgajer.github.io/geosmooth/reference/normalize.density.html) | Convert a numeric field or supported fitted object to probability mass, with normalization diagnostics. |
| [`fit.subject.od()`](https://pgajer.github.io/geosmooth/reference/fit.subject.od.html) | Convert repeated visit indices into counts and fit a subject-occupation distribution. |

```{r occupation}
support <- matrix(seq(0, 1, length.out = 6), ncol = 1)
occupation <- fit.subject.od(
  support, subject.index = c(1L, 3L, 3L, 6L), method = "empirical"
)
occupation$rho
sum(occupation$rho)
```

Here the third support point receives twice as many visits as the first or
sixth. `rho` is a probability **mass** vector over support points, summing to
one; it is not automatically a continuous density per unit volume.
[`normalize.density()`](https://pgajer.github.io/geosmooth/reference/normalize.density.html) is an explicit conversion step, not another regression
fit. Inspect its clipping and normalization accounting when the source field
contains negative values.

[`fit.subject.od()`](https://pgajer.github.io/geosmooth/reference/fit.subject.od.html) offers more methods than [`fit.density()`](https://pgajer.github.io/geosmooth/reference/fit.density.html): it can fit ordinary
smoothers and normalize their output. Visit-level CV holds out visits and uses
`visit.foldid`; ordinary regression CV holds out support rows and uses
`foldid` (or the method-specific spelling). Keep these validation units distinct.

### Experimental transport methods

| Function | Purpose |
|---|---|
| [`pttf.geometry()`](https://pgajer.github.io/geosmooth/reference/pttf.geometry.html) | Build local tangent frames and their transport across a graph. |
| [`pttf.operator()`](https://pgajer.github.io/geosmooth/reference/pttf.operator.html) | Construct parallel-transport trend-filtering differences from that geometry. |
| [`pttf.operator.filter.rows()`](https://pgajer.github.io/geosmooth/reference/pttf.operator.filter.rows.html) | Select operator rows while maintaining provenance and related operator data. |
| [`fit.pttf.trend.filtering()`](https://pgajer.github.io/geosmooth/reference/fit.pttf.trend.filtering.html) | Fit a parallel-transport trend filter with an absolute-value or quadratic penalty. |
| [`transported.graph.hessian.operator()`](https://pgajer.github.io/geosmooth/reference/transported.graph.hessian.operator.html) | Build a separate experimental graph Hessian diagnostic with several direction-matching rules. |

Parallel transport aligns local tangent directions before comparing
derivatives. These experimental interfaces expose geometry, operator, and
solver diagnostics deliberately. The transported graph Hessian constructor is
a separate diagnostic workflow; it does not fit a response and is not an
interchangeable spelling of [`pttf.operator()`](https://pgajer.github.io/geosmooth/reference/pttf.operator.html).

### Synthetic datasets: specifications, responses, and reproducibility

| Function | Purpose |
|---|---|
| [`synthetic.spec()`](https://pgajer.github.io/geosmooth/reference/synthetic.spec.html) | Combine geometry, sampling, truth, and response descriptors without drawing data. |
| [`materialize.synthetic()`](https://pgajer.github.io/geosmooth/reference/materialize.synthetic.html) | Draw a dataset from a specification with an explicit seed and random-number policy. |
| [`materialize.synthetic.instance()`](https://pgajer.github.io/geosmooth/reference/materialize.synthetic.instance.html) | Reproduce a registered frozen instance and optionally validate its checksum. |
| [`synthetic.registry.ids()`](https://pgajer.github.io/geosmooth/reference/synthetic.registry.ids.html) | List maintained recipe identifiers. |
| [`synthetic.registry.spec()`](https://pgajer.github.io/geosmooth/reference/synthetic.registry.spec.html) | Look up a recipe specification; some recipe families allow parameter overrides. |
| [`synthetic.registry.seed()`](https://pgajer.github.io/geosmooth/reference/synthetic.registry.seed.html) | Reproduce legacy SSRHE recipe seeds; this is not a seed generator for every registry family. |
| [`validate.synthetic.dataset()`](https://pgajer.github.io/geosmooth/reference/validate.synthetic.dataset.html) | Check the structure and consistency of a materialized dataset. |
| [`compare.synthetic.dataset()`](https://pgajer.github.io/geosmooth/reference/compare.synthetic.dataset.html) | Compare two datasets with explicit numerical tolerances. |
| [`synthetic.dataset.checksum()`](https://pgajer.github.io/geosmooth/reference/synthetic.dataset.checksum.html) | Compute a content checksum for reproducibility checks. |
| [`synthetic.sampling.stratified()`](https://pgajer.github.io/geosmooth/reference/synthetic.sampling.stratified.html) | Describe allocation between the supported synthetic strata. |
| [`synthetic.truth.named()`](https://pgajer.github.io/geosmooth/reference/synthetic.truth.named.html) | Select a named truth specification with optional parameters. |
| [`synthetic.truth.polynomial()`](https://pgajer.github.io/geosmooth/reference/synthetic.truth.polynomial.html) | Specify polynomial truth in the chosen evaluation coordinates. |
| [`synthetic.truth.gaussian.mixture()`](https://pgajer.github.io/geosmooth/reference/synthetic.truth.gaussian.mixture.html) | Specify Gaussian-mixture truth using centers, scales, and weights. |
| [`synthetic.truth.occupation.mixture()`](https://pgajer.github.io/geosmooth/reference/synthetic.truth.occupation.mixture.html) | Specify occupation-mixture truth with covariance and probability controls. |
| [`synthetic.truth.logit()`](https://pgajer.github.io/geosmooth/reference/synthetic.truth.logit.html) | Specify a logit-based truth with prevalence and clipping controls. |
| [`synthetic.response.gaussian()`](https://pgajer.github.io/geosmooth/reference/synthetic.response.gaussian.html) | Specify Gaussian response noise with a supplied standard deviation. |
| [`synthetic.response.heteroskedastic.gaussian()`](https://pgajer.github.io/geosmooth/reference/synthetic.response.heteroskedastic.gaussian.html) | Specify Gaussian noise whose scale depends on truth. |
| [`synthetic.response.clustered.gaussian()`](https://pgajer.github.io/geosmooth/reference/synthetic.response.clustered.gaussian.html) | Specify within-cluster correlated Gaussian noise. |
| [`synthetic.response.laplace.outlier()`](https://pgajer.github.io/geosmooth/reference/synthetic.response.laplace.outlier.html) | Specify Laplace noise and an outlier component. |
| [`synthetic.response.bernoulli()`](https://pgajer.github.io/geosmooth/reference/synthetic.response.bernoulli.html) | Specify binary response draws and optional positive-count requirements. |

The constructors create specifications; [`materialize.synthetic()`](https://pgajer.github.io/geosmooth/reference/materialize.synthetic.html) performs
random draws. Use `print()`, `plot()`, and `as.data.frame()` on a resulting
`synthetic_dataset`. For complete examples, read
`vignette("synthetic-datasets", package = "geosmooth")`. Use the named-stream
random-number policy for new work; historical seed and algorithm controls
preserve frozen examples and should not be changed casually.

### Geometry constructors and helpers supplied by dgraphs

These names are available from `dgraphs`, and are no longer exported by
`geosmooth`. Replace former `geosmooth::` calls with `dgraphs::`. See, for example,
`help("synthetic.circle", package = "dgraphs")`.

| Function | Purpose |
|---|---|
| `dgraphs::synthetic.circle()` | Describe a circle or circular arc and its embedding. |
| `dgraphs::synthetic.helix()` | Describe a helical curve. |
| `dgraphs::synthetic.trefoil()` | Describe a trefoil curve. |
| `dgraphs::synthetic.torus.patch()` | Describe a patch of a torus. |
| `dgraphs::synthetic.sphere.cap()` | Describe a spherical cap. |
| `dgraphs::synthetic.quadform()` | Describe a flat or quadratic geometry with intrinsic and ambient dimensions. |
| `dgraphs::synthetic.simplex()` | Describe a simplex with a given number of parts. |
| `dgraphs::synthetic.stratified()` | Combine strata and optional junctions. |
| `dgraphs::synthetic.stratum.point()` | Describe a point stratum. |
| `dgraphs::synthetic.stratum.segment()` | Describe a line-segment stratum. |
| `dgraphs::synthetic.stratum.rectangle()` | Describe a rectangular stratum. |
| `dgraphs::synthetic.point.line.junction()` | Describe a point-and-line junction geometry. |
| `dgraphs::embed.synthetic.geometry()` | Map latent coordinates to the ambient space. |
| `dgraphs::edge.lengths.synthetic.geometry()` | Evaluate geometry-aware lengths between latent endpoints. |
| `dgraphs::quadform.gradient()` | Evaluate the gradient of a quadratic geometry. |
| `dgraphs::quadform.metric()` | Evaluate its induced metric. |

### Sampling constructors supplied by dgraphs

| Function | Purpose |
|---|---|
| `dgraphs::synthetic.sampling.uniform.interval()` | Describe uniform draws on an interval. |
| `dgraphs::synthetic.sampling.uniform.box()` | Describe uniform draws in a box. |
| `dgraphs::synthetic.sampling.uniform.rectangle()` | Describe uniform draws in a rectangle with its specified draw algorithm. |
| `dgraphs::synthetic.sampling.uniform.disk()` | Describe uniform draws in a disk. |
| `dgraphs::synthetic.sampling.grid.interval()` | Describe an interval grid with an endpoint convention. |
| `dgraphs::synthetic.sampling.gapped.uniform()` | Describe sampling across separated intervals. |
| `dgraphs::synthetic.sampling.truncated.normal()` | Describe bounded Gaussian draws. |
| `dgraphs::synthetic.sampling.clustered()` | Describe clustered observations around random centers. |
| `dgraphs::synthetic.sampling.dirichlet.zeros()` | Describe Dirichlet sampling with structural zeros. |

Similar geometric domains do not imply identical random-number sequences:
these descriptors also preserve ordering and algorithm choices.

## Finding detailed help

Function links in this guide open the rendered online reference. When working
without an internet connection, the same help is installed with the package:
use `help("fit.lps", package = "geosmooth")`, replacing `fit.lps` with the
function name. Start at `help("geosmooth-package", package = "geosmooth")`
for clickable task links in installed help, or use
`browseVignettes("geosmooth")` for the installed HTML guides.

The [synthetic datasets guide](synthetic-datasets.html) develops reproducible
geometry, sampling, truth, and response specifications in more detail.


```{r help-navigation, eval=FALSE}
help(package = "geosmooth")
help("geosmooth-package", package = "geosmooth")
help("fit.lps", package = "geosmooth")
help("predict.lps", package = "geosmooth")
vignette("function-guide", package = "geosmooth")
vignette("synthetic-datasets", package = "geosmooth")
```

The package's help pages give exact arguments, defaults, return values, and
examples. Use this guide to find the appropriate family, then read the chosen
function's help before configuring its advanced controls.
