---
title: "Synthetic datasets for geometric smoothing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synthetic datasets for geometric smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

For the complete task and function catalog, see
[Finding your way around geosmooth](function-guide.html).

`geosmooth` represents a synthetic dataset as four serializable components:
geometry, sampling, truth, and response. [`synthetic.spec()`](https://pgajer.github.io/geosmooth/reference/synthetic.spec.html) combines those
components without drawing data. [`materialize.synthetic()`](https://pgajer.github.io/geosmooth/reference/materialize.synthetic.html) performs all random
work under an explicit seed and RNG policy and returns a validated
`synthetic_dataset`.

```{r circle}
library(geosmooth)

circle.spec <- synthetic.spec(
  geometry = dgraphs::synthetic.circle(radius = 1, ambient.dim = 2),
  sampling = dgraphs::synthetic.sampling.grid.interval(
    0, 2 * pi, endpoints = "exclude.lower"),
  truth = synthetic.truth.polynomial(c(b0 = 0)),
  response = synthetic.response.gaussian(sd = 0.05)
)
circle.data <- materialize.synthetic(circle.spec, n = 80, seed = 2026)
circle.data
```

The circle has one latent coordinate (angle) and two observed coordinates
(the embedding). The constant truth is intentional: this recipe isolates
response noise around zero, rather than testing recovery of a varying signal.

```{r circle-views, fig.width=8, fig.height=4, out.width="100%", fig.cap="Left: the circle in its two observed coordinates. Right: responses versus latent angle, with the constant zero truth shown by a dashed line. These views describe the same 80 observations.", fig.alt="A circle of 80 points beside noisy responses scattered around a dashed zero line over angles from zero to two pi."}
old.par <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(circle.data, view = "geometry", col = "#0072B2", asp = 1,
     main = "Observed geometry")
plot(circle.data, view = "response", col = "grey40", xlab = "Angle (radians)",
     main = "Response and truth")
par(old.par)
```

Quadforms use separate intrinsic and ambient dimensions. A flat geometry has no
quadratic forms; a paraboloid or saddle supplies one or more symmetric forms.

```{r quadform}
surface.spec <- synthetic.spec(
  geometry = dgraphs::synthetic.quadform(
    intrinsic.dim = 2,
    ambient.dim = 5,
    forms = list(diag(c(1, -1))),
    frame = "random.orthonormal"
  ),
  sampling = dgraphs::synthetic.sampling.uniform.box(-1, 1),
  truth = synthetic.truth.polynomial(c(b0 = 0, b1 = 1, b22 = -0.5)),
  response = synthetic.response.gaussian(sd = 0.1)
)
surface.data <- materialize.synthetic(surface.spec, n = 100, seed = 17)
dim(surface.data$latent)
dim(surface.data$predictors)
head(as.data.frame(surface.data))
```

The surface has two latent coordinates and five observed coordinates. The
random orthonormal frame embeds the surface in five dimensions. Viewing two
coordinates is a projection: points that appear close may be separated in
coordinates omitted from the display.

```{r surface-view, fig.width=6, fig.height=4.5, out.width="80%", fig.cap="Two-coordinate projection of a surface embedded in five dimensions. Point colors indicate the known, nonconstant conditional mean; the key uses its actual numerical range.", fig.alt="A projected surface point cloud colored by its known conditional mean, with a numeric color key."}
plot(surface.data, view = "geometry", color = "truth", main = "Projected surface")
```

This second recipe has a varying polynomial truth. A smoother receives the
observed predictors and noisy response, not the latent coordinates or truth:

```{r surface-fit}
surface.fit <- fit.lps(surface.data$predictors, surface.data$response,
  foldid = rep(1:5, length.out = surface.data$n),
  support.grid = c(20L, 30L), degree.grid = 1L,
  kernel.grid = "gaussian", coordinate.method = "local.pca", chart.dim = 2L)
surface.fit$selected
head(data.frame(response = surface.data$response, fitted = fitted(surface.fit),
                truth = surface.data$truth))
```

The fitted values are on the observed rows. Truth is available for this
synthetic demonstration only; selecting settings and reporting their CV score
does not supply an independent test of prediction accuracy. The
[first-fit guide](function-guide.html#a-first-regression-fit) explains that distinction.

Maintained benchmark recipes are discoverable through
[`synthetic.registry.ids()`](https://pgajer.github.io/geosmooth/reference/synthetic.registry.ids.html). G1--G7 remain compatibility identifiers; the
S01--S16 shape identifiers and V1--V3 sampling/noise variants preserve the shared
SSRHE regression suite. For example, S01.V1 means two separated narrow equal
signal peaks with uniform sampling and Gaussian response noise. G1 is an ambient
polynomial example; G2 is a polynomial on a flat embedded geometry.
Ordinary datasets bind their full specification hash into `dataset.id`.

```{r registry}
ids <- c("G1", "G2", "S01.V1")
do.call(rbind, lapply(ids, function(id) {
  spec <- synthetic.registry.spec(id)
  data.frame(id, geometry = spec$geometry.spec$family,
             truth = spec$truth.spec$parameters$estimand)
}))
registered <- synthetic.registry.spec("S01.V1")
registered.data <- materialize.synthetic(
  registered, n = 160,
  seed = synthetic.registry.seed("S01.V1", replicate = 1),
  rng.policy = "legacy"
)
synthetic.dataset.checksum(registered.data)
```

Use `rng.policy = "named.stream.v1"` for new work. The `"legacy"` policy exists
for frozen scientific parity and pins historical RNG behavior. Predictor noise,
response noise, and finite-design truth normalization are distinct contracts;
they should not be substituted for one another.
