---
title: "LPS Tier 2 — binary-path correctness gates: E2.14 (separation robustness) and E2.12 (selection-metric consistency)"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
  pdf_document:
    toc: true
    toc_depth: 2
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10.5,
  fig.height = 6.2,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "ragg_png"
)
library(ggplot2)

# LaTeX auto-numbers kable captions ("Table N: ..."); the HTML build keeps
# the manual visible numbering required by the HTML style guide.
tab.cap <- function(n, text) {
  if (knitr::is_latex_output()) text else paste0("Table ", n, ". ", text)
}

# Report-specific inputs live beside the rendered report; the gate/study
# evidence CSVs stay at their canonical package-level reports/ location
# (they are wired into the Tier-2 harness and the frozen handoffs).
inputs.dir <- file.path("dev", "methods", "lps", "reports", "tier2",
                        "binary_hygiene", "2026-06-11", "inputs")
fixture14 <- read.csv(file.path(inputs.dir, "e2_14_fixture.csv"))
prefix14  <- read.csv("reports/e2_14_prefix_newton_overshoot.csv")
traces14  <- read.csv(file.path(inputs.dir, "e2_14_deviance_traces.csv"))
summary14 <- read.csv(file.path(inputs.dir, "e2_14_solver_summary.csv"))
fallback14 <- read.csv(file.path(inputs.dir, "e2_14_fallback_telemetry.csv"))
fixture12a <- read.csv(file.path(inputs.dir, "e2_12_fixture_a.csv"))
sel12a    <- read.csv(file.path(inputs.dir, "e2_12a_selection_metric.csv"))
contrib12b <- read.csv(file.path(inputs.dir, "e2_12b_contributions.csv"))
dom12b    <- read.csv(file.path(inputs.dir, "e2_12b_clip_dominance.csv"))
acct12b   <- read.csv(file.path(inputs.dir, "e2_12b_fit_accounting.csv"))
crossclip <- read.csv("reports/e2_12_crossclip_scores.csv")
verdict   <- read.csv("reports/e2_12_crossclip_stability_verdict.csv")
manifest  <- readLines(file.path(inputs.dir,
                                 "tier2_report_inputs_manifest.txt"))

mf <- function(key) {
  ln <- grep(paste0("^", key, ":"), manifest, value = TRUE)[1]
  trimws(sub(paste0("^", key, ":"), "", ln))
}
build.time <- Sys.getenv("BUILD_DATETIME",
  format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "America/New_York"))

col.pre  <- "#B23A48"   # pre-fix / raw metric
col.post <- "#2C6E91"   # post-fix / deployed metric
col.flag <- "#D7263D"   # highlighted (flipped) observation
```

<div style="border:1px solid #ccc; background:#f8f8f8; padding:0.6em 1em; font-size:90%;">
**Report build:** `r build.time`\
**Result generation:** `r mf("results_generated_eastern")` (UTC `r mf("results_generated_utc")`)\
**Source commit:** `` `r substr(mf("git_head"), 1, 12)` `` (branch `codex/geosmooth-t2-binary-hygiene`)\
**Report source (repo-relative):** `dev/methods/lps/reports/tier2/binary_hygiene/2026-06-11/`\
`scripts/lps_tier2_binary_hygiene_report.Rmd`\
**Input manifest:** [`inputs/tier2_report_inputs_manifest.txt`](inputs/tier2_report_inputs_manifest.txt)\
**Bundle cross-check:** `r mf("bundle_crosscheck")`
</div>

All numbers shown are read from precomputed, committed CSV artifacts — the
report-specific inputs in `inputs/` beside this report (regenerated
deterministically by the adjacent `scripts/export_tier2_report_inputs.R`
from the gated fixture constructors) and the gate/study evidence at the
package-level `reports/` directory; this report did not rerun the
experiments. The acceptance evidence proper lives in the execution bundles
`audit_artifacts/tier2_20260611T075822Z` (E2.14, commit `75c1788`) and
`audit_artifacts/tier2_20260611T084642Z` (E2.12 + E2.14 rerun, commit
`3eaf9cf`), whose regenerated aggregates the input manifest cross-checks.

# Purpose

`fit.lps()` fits local weighted polynomial regression in local charts and
selects its configuration (support size, polynomial degree, kernel) by
cross-validation. Its binary path has two modes: `bernoulli` (local least
squares on 0/1 labels; deployed probabilities are the predictions clipped to
$[0,1]$) and `binomial` (local weighted logistic fits solved by iteratively
reweighted least squares, IRLS; selection by cross-validated log loss).

This report documents the results of the two Tier-2 correctness gates that
landed on branch `codex/geosmooth-t2-binary-hygiene`:

1. **E2.14 — separation robustness.** Is the local logistic IRLS monotone
   and bounded on (near-)separable supports, and does *exact* separation end
   in a documented, telemetered fallback rather than an unbounded loop or
   `NaN`? The fix under test is deviance **step-halving** inside the IRLS
   loop.
2. **E2.12 — selection-metric consistency and log-loss clipping.**
   (a) Does bernoulli-mode selection score the **deployed** (clipped)
   metric rather than raw, unclipped predictions — and can the difference
   change which candidate is selected? (b) Is the log-loss probability
   truncation pinned at $10^{-6}$, and what instability did the previous
   $10^{-15}$ truncation permit? A companion **study** (reported, not
   gated) asks whether the selected candidate is stable across truncation
   levels $\{10^{-6}, 10^{-3}\}$.

Both gates are deterministic `testthat` batteries on constructed fixtures
(no replication), so all results below are single realized values; no
uncertainty intervals apply. The committed gates, both passing on the
committed tree alongside the full Tier-0 battery inside the bundles named
above, are:

- `tests/testthat/test-lps-binary-separation.R` (E2.14, 43 assertions);
- `tests/testthat/test-lps-binary-metric-consistency.R` (E2.12, 43
  assertions including the audit-required backend-resolution coverage).

# E2.14 — the constructed separation fixtures

A local logistic solve receives a small neighborhood: 1-D local coordinates
$z_i$ centered at the evaluation point, labels $y_i \in \{0,1\}$, and
gaussian kernel weights $w_i$ that decay with $|z_i|$. The gate uses one
support shape in two label configurations (the frozen plan's DGP: $y =
\mathbf{1}\{z>0\}$, optionally one flipped label, weights from the gaussian
kernel):

- **Near-separable arm:** ten equispaced negatives $z \in \{-0.20, -0.18,
  \dots, -0.02\}$ plus one far positive at $z = 6$, with the label of the
  *second* point ($z = -0.18$) flipped to $1$. Two classes, not linearly
  separable: the flipped positive sits inside the negative cluster.
- **Exactly separable arm:** the same support without the flip — a
  threshold at any $z \in (-0.02, 6)$ separates the classes perfectly, so
  the unpenalized logistic maximum-likelihood estimate does not exist
  (coefficients diverge).

The flip *position* is load-bearing: flipping an edge-of-cluster point
produces the plain-Newton overshoot shown in Figure 2, while a mid-cluster
flip leaves plain Newton monotone and would make the fix undetectable.

```{r fig1-fixture, fig.height=4.6}
fx <- rbind(
  transform(fixture14, arm = "near-separable", y = y.near.separable),
  transform(fixture14, arm = "exactly separable", y = y.exactly.separable)
)
fx$arm <- factor(fx$arm, levels = c("near-separable", "exactly separable"))
fx$region <- factor(
  ifelse(fx$z < 0, "negative cluster (z < 0)", "far positive (z = 6)"),
  levels = c("negative cluster (z < 0)", "far positive (z = 6)"))
fx$highlight <- fx$flipped & fx$arm == "near-separable"
ggplot(fx, aes(x = z, y = y)) +
  geom_point(aes(size = weight, color = highlight), alpha = 0.85) +
  facet_grid(arm ~ region, scales = "free_x", space = "free_x") +
  scale_color_manual(values = c(`FALSE` = "grey35", `TRUE` = col.flag),
                     labels = c(`FALSE` = "as constructed",
                                `TRUE` = "flipped label"),
                     name = NULL) +
  scale_size_continuous(name = "kernel weight", range = c(2, 5),
                        breaks = c(0.7, 0.9, 1.0)) +
  scale_y_continuous(breaks = c(0, 1), limits = c(-0.25, 1.25)) +
  labs(x = "local coordinate z", y = "label y") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom",
        strip.text.y = element_text(size = 10))
```

**Figure 1.** The E2.14 fixture in its two label configurations (rows:
near-separable = one flipped label; exactly separable = no flip). The
x-axis is split (columns, independent scales) because the far positive at
$z = 6$ would otherwise compress the negative cluster into a single blot.
Point size encodes the gaussian kernel weight: cluster points carry weight
$\approx 1.0$, the far positive $\approx 0.61$. The red point in the top row
is the single flipped label at $z = -0.18$ — a high-weight, wrong-side
observation. Removing the flip (bottom row) makes the support exactly
separable.

# E2.14 — deviance trajectories: plain Newton vs step-halving

The solver minimizes the weighted binomial deviance

$$
D(\beta) \;=\; -2 \sum_i w_i \Big( y_i \log \mu_i(\beta)
              + (1-y_i)\log\big(1-\mu_i(\beta)\big) \Big),
\qquad
\mu_i(\beta) = \operatorname{expit}\!\big(\tilde\eta_i\big),
$$

where $\tilde\eta_i$ is the linear predictor $\eta_i = x_i^\top\beta$ clamped
to $[-35, 35]$ (the same clamp the IRLS update uses, which keeps $D$ finite
for every finite iterate). Plain IRLS accepts every Newton update
unconditionally. The E2.14 fix accepts a candidate $\beta_{\text{new}}$ only
if $D(\beta_{\text{new}}) \le D(\beta_{\text{cur}}) + 10^{-8}$, halving the
step toward $\beta_{\text{cur}}$ up to 30 times otherwise; an exhausted
halving budget ends the solve as non-convergent. The gate asserts on the
whole recorded trajectory, not only its endpoint, so oscillation cannot
hide.

```{r fig2-trajectories, fig.height=5.6}
near.pre  <- transform(prefix14, algorithm = "pre-fix plain Newton")
near.post <- transform(
  subset(traces14, arm == "near.separable")[, c("iteration", "deviance")],
  algorithm = "post-fix step-halving")
near <- rbind(near.pre, near.post)
near$panel <- "near-separable arm"
exact <- transform(
  subset(traces14, arm == "exactly.separable")[, c("iteration", "deviance")],
  algorithm = "both algorithms (identical here)")
exact$panel <- "exactly separable arm"
both <- rbind(near, exact)
both$panel <- factor(both$panel,
  levels = c("near-separable arm", "exactly separable arm"))
both$algorithm <- factor(both$algorithm,
  levels = c("pre-fix plain Newton", "post-fix step-halving",
             "both algorithms (identical here)"))
end.post <- transform(subset(near.post, iteration == max(iteration)),
                      panel = factor("near-separable arm",
                                     levels = levels(both$panel)))
ggplot(both, aes(iteration, deviance, color = algorithm)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  geom_point(data = end.post,
             shape = 21, size = 3.4, stroke = 1.1, fill = "white") +
  facet_wrap(~panel, nrow = 1, scales = "free") +
  scale_y_log10() +
  scale_color_manual(values = c(`pre-fix plain Newton` = col.pre,
                                `post-fix step-halving` = col.post,
                                `both algorithms (identical here)` = "grey30"),
                     name = NULL) +
  labs(x = "accepted IRLS iteration (0 = initialization)",
       y = "weighted binomial deviance (log scale)") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

**Figure 2.** Per-iteration deviance of the local logistic solve on the two
fixture arms (log scale). *Left:* on the near-separable arm, pre-fix plain
Newton (red; replicated update equations of the pre-change solver, committed
trajectory in `reports/e2_14_prefix_newton_overshoot.csv`) overshoots at
iteration 2 — the deviance jumps from $6.9$ to $\approx 496$, a $+489$
increase — then oscillates between $\approx 6.5$ and $\approx 9$ and never
converges within the 50-iteration cap. Post-fix step-halving (blue) needs a
single halving at that same step, descends monotonically (largest per-step
increase $1.8 \times 10^{-15}$, within the $10^{-8}$ slack), and converges at
iteration 9 (open circle) with finite coefficients and a center prediction
$\hat p = 0.125 \in (0,1)$. *Right:* on the exactly separable arm both
algorithms coincide (no halving triggers): the deviance decreases
monotonically toward zero as the coefficients run off to infinity, and the
solve stops at the 50-iteration cap as designed.

The realized endpoint quantities behind Figure 2:

```{r table1}
t1 <- summary14
t1$max.deviance.increase <- sprintf("%.2e", t1$max.deviance.increase)
t1$prediction <- ifelse(is.na(t1$prediction), "--",
                        sprintf("%.4f", t1$prediction))
knitr::kable(
  t1, col.names = c("arm", "status", "converged", "iterations",
                    "step halvings", "max dev. increase", "p-hat (center)"),
  caption = tab.cap(1, paste(
    "Post-fix solver summary per fixture arm.",
    "Columns: solver status string; whether the coefficient-step",
    "criterion (1e-7) was met within the 50-iteration",
    "cap; accepted iterations; total step halvings; the",
    "largest per-step deviance increase along the accepted",
    "trajectory (the gate requires it below 1e-8); the fitted center",
    "probability ('--' = no converged fit is returned; the",
    "fallback layer supplies the prediction, Table 2).")))
```

On non-convergence the fitting layer falls back deterministically and
records it: under `unstable.action = "mean"` the prediction is the local
weighted event rate $\bar y_w = \sum_i w_i y_i / \sum_i w_i$; under
`unstable.action = "na"` it is `NA` (the sanctioned guarded output). The
exactly separable arm exercises both paths:

```{r table2}
t2 <- fallback14[, c("unstable.action", "fitted", "attempted", "converged",
                     "fallback.path.count", "event.rate.fallback.count",
                     "na.failure.count")]
t2$fitted <- ifelse(is.na(t2$fitted), "NA", sprintf("%.6f", t2$fitted))
knitr::kable(
  t2, col.names = c("unstable.action", "returned value", "attempted",
                    "converged", "fallback", "event-rate fallback",
                    "NA failure"),
  caption = tab.cap(2, paste(
    "Fallback behavior and telemetry on the exactly",
    "separable arm (one solve attempt each). 'returned value'",
    "is the fitted probability the caller receives: the",
    "weighted event rate 0.063151 under \"mean\", NA under",
    "\"na\". The last four columns are the recorded telemetry",
    "counters (logistic.diagnostics): every fallback is",
    "counted, split into event-rate substitutions and NA",
    "returns. No NaN and no unbounded loop occurs.")))
```

E2.14 also pins that nothing changed where nothing needed to change: when
every full Newton step already satisfies the deviance slack, the halving
check accepts the identical iterate, so well-behaved solves are numerically
identical to the pre-change solver (asserted in the gate against an
independent plain-IRLS replication, and corroborated by the Tier-0 E0.6
battery statistics being bit-identical to the unmodified base commit).

# E2.12 — the constructed G6 fixture

Both E2.12 sub-items use the plan's binary-surface DGP **G6** with one named
override: a *sharp* log-odds cliff in place of the default smooth surface,

$$
\eta(x) = 6\,\tanh(15\,x_1), \qquad
p(x) = \min\!\big(\max(\operatorname{expit}(\eta(x)),\,0.05),\,0.95\big),
\qquad y \sim \mathrm{Bernoulli}(p),
$$

on $x \in [-1,1]^2$ with $n = 400$ (prevalence target $0.5$ by symmetry; the
$[0.05, 0.95]$ probability clamp is part of the G6 construction). The cliff
at $x_1 = 0$ is what forces flexible local fits to overshoot $[0,1]$ on the
raw scale — the behavior sub-item (a) needs. Sub-item (b) uses a second draw
(seed 7001) of the same surface with one label deliberately flipped, fit in
`binomial` mode.

```{r fig3-g6, fig.height=4.4}
curve.df <- fixture12a[order(fixture12a$x1), ]
ggplot() +
  geom_line(data = curve.df, aes(x1, p), linewidth = 0.9,
            color = col.post) +
  geom_jitter(data = fixture12a, aes(x1, y), width = 0, height = 0.04,
              alpha = 0.35, size = 1.1, color = "grey25") +
  scale_y_continuous(breaks = c(0, 0.05, 0.5, 0.95, 1)) +
  labs(x = expression(x[1]),
       y = "p(x) (line)  /  drawn label y (points)") +
  theme_bw(base_size = 13)
```

**Figure 3.** The E2.12(a) fixture (seed 2121): the true success probability
$p(x)$ (blue line; it depends on $x_1$ only) and the $n = 400$ drawn labels
(grey points, vertically jittered). The surface is a two-plateau cliff —
$p = 0.05$ for $x_1 \lesssim -0.15$, $p = 0.95$ for $x_1 \gtrsim 0.15$ —
so local polynomial fits near the transition extrapolate outside $[0,1]$.
The E2.12(b) fixture is a second draw (seed 7001) of this same surface; its
deliberately flipped observation sits deep in the left plateau at
$x_1 = -0.702$, where every 60-point neighborhood is all-zero.

# E2.12(a) — selection scores the deployed clipped metric

For a candidate configuration $j$ with out-of-fold (cross-validation)
predictions $\hat p_{ij}$, the two scores at issue are

$$
\text{raw Brier}_j = \frac{1}{n}\sum_i \big(y_i - \hat p_{ij}\big)^2,
\qquad
\text{deployed Brier}_j = \frac{1}{n}\sum_i
\big(y_i - \operatorname{clip}_{[0,1]}(\hat p_{ij})\big)^2,
$$

where $\operatorname{clip}_{[0,1]}$ is exactly the clamp applied to the
probabilities the fitted model actually deploys
(`fitted.values`, `predict()`). Pre-fix, bernoulli selection minimized the
raw score while deployment clipped — a train/deploy metric mismatch.
Post-fix, the selection column `cv.brier.observed` *is* the deployed score
(with `Inf` for any candidate having a non-finite prediction, preserving
the old `NA`-handling semantics).

Fit-status accounting for this fixture: all $4 \times 400 = 1600$
out-of-fold predictions are finite (no `NA` candidates, nothing excluded);
263 of them (16.4%) fall outside $[0,1]$, concentrated in the two
degree-2 candidates — the clip is doing real work.

```{r fig4-flip, fig.height=4.6}
sel <- sel12a
sel$label <- sprintf("support %d, degree %d", sel$support.size, sel$degree)
sel$label <- factor(sel$label,
                    levels = sel$label[order(-sel$deployed.clipped.brier)])
long <- rbind(
  data.frame(label = sel$label, metric = "raw Brier (pre-fix selection)",
             score = sel$raw.brier),
  data.frame(label = sel$label, metric = "deployed clipped Brier (post-fix)",
             score = sel$deployed.clipped.brier)
)
winners <- rbind(
  data.frame(label = sel$label[which.min(sel$raw.brier)],
             metric = "raw Brier (pre-fix selection)",
             score = min(sel$raw.brier)),
  data.frame(label = sel$label[which.min(sel$deployed.clipped.brier)],
             metric = "deployed clipped Brier (post-fix)",
             score = min(sel$deployed.clipped.brier))
)
ggplot(long, aes(score, label, color = metric, shape = metric)) +
  geom_line(aes(group = label), color = "grey60", linewidth = 0.6) +
  geom_point(data = subset(long, metric == "raw Brier (pre-fix selection)"),
             size = 4.6, stroke = 1.2) +
  geom_point(data = subset(long,
                           metric == "deployed clipped Brier (post-fix)"),
             size = 2.6) +
  geom_point(data = winners, shape = 21, size = 7.5, stroke = 1.2,
             fill = NA, show.legend = FALSE) +
  scale_x_log10(breaks = c(0.07, 0.1, 0.2, 0.3, 0.5, 0.9)) +
  scale_color_manual(values = c(`raw Brier (pre-fix selection)` = col.pre,
                                `deployed clipped Brier (post-fix)` = col.post),
                     name = NULL) +
  scale_shape_manual(values = c(`raw Brier (pre-fix selection)` = 5,
                                `deployed clipped Brier (post-fix)` = 16),
                     name = NULL) +
  labs(x = "cross-validated Brier score (log scale; lower is better)",
       y = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

**Figure 4.** Per-candidate selection scores on the E2.12(a) fixture under
the two metrics (one row per candidate; open red diamonds = raw Brier,
filled blue dots = deployed clipped Brier; grey segments connect a
candidate's two scores; log x-axis; large circles mark each metric's
minimizer). Where a candidate's predictions never leave $[0,1]$ the two
scores coincide exactly and the blue dot sits inside the red diamond.
Clipping therefore moves nothing for the in-range candidates but rescues the
degree-2 candidates whose raw predictions overshoot — most dramatically
(support 8, degree 2), whose score falls from $0.883$ to $0.110$. The two
metrics disagree on the winner: the raw score (the pre-fix rule) picks
(support 8, degree 0) while the deployed score picks (support 60, degree 2),
with cross-candidate margins of $5.8 \times 10^{-4}$ (raw) and
$2.1 \times 10^{-3}$ (deployed) — both more than $100\times$ the contract's
$10^{-6}$ comparison tolerance, so the flip is real rather than a tie. The
post-fix fit deploys the clipped winner.

```{r table3}
t3 <- sel12a[order(-sel12a$deployed.clipped.brier),
             c("support.size", "degree", "raw.brier",
               "cv.brier.observed", "n.pred.below.0", "n.pred.above.1",
               "selected")]
t3$raw.brier <- sprintf("%.6f", t3$raw.brier)
t3$cv.brier.observed <- sprintf("%.6f", t3$cv.brier.observed)
knitr::kable(
  t3, row.names = FALSE,
  col.names = c("support", "degree", "raw Brier", "deployed Brier",
                "n pred < 0", "n pred > 1", "selected"),
  caption = tab.cap(3, paste(
    "The four candidates on the E2.12(a) fixture.",
    "'raw Brier' is the mean squared error of the unclipped",
    "out-of-fold predictions (the pre-fix selection score);",
    "'deployed Brier' is the same on [0,1]-clipped",
    "predictions and is the post-fix selection column",
    "cv.brier.observed; the two count columns give each",
    "candidate's out-of-fold predictions outside [0,1] (of",
    "400); 'selected' marks the candidate the post-fix fit",
    "returns. The gate additionally verifies the selection",
    "column equals the deployed metric recomputed from the",
    "stored CV predictions (max residual 1.4e-17, contract",
    "tolerance 1e-6).")))
```

# E2.12(b) — the log-loss clip: one point owned the score

Binomial-mode selection minimizes the cross-validated log loss with
probability truncation $\varepsilon$:

$$
\mathrm{LL}_\varepsilon(j) = -\frac{1}{n}\sum_i \Big(
 y_i \log \operatorname{clip}_{[\varepsilon,\,1-\varepsilon]}(\hat p_{ij})
 + (1-y_i)\log\big(1-\operatorname{clip}_{[\varepsilon,\,1-\varepsilon]}(\hat p_{ij})\big)\Big).
$$

A confidently wrong held-out point ($\hat p \approx 0$, $y = 1$) contributes
$-\log\varepsilon$ — $34.54$ at the previous $\varepsilon = 10^{-15}$ versus
$13.82$ at the pinned $\varepsilon = 10^{-6}$. On the (b) fixture the
deliberately flipped observation's out-of-fold prediction is exactly $0$
under both candidates: its 60-point neighborhoods are all-zero, so the local
logistic solve lands in the event-rate fallback, whose weighted mean of
zeros is $0$.

Fit-status accounting: the binomial cross-validation ran 800 local logistic
solves (400 points $\times$ 2 candidates); 785 converged (98.1%) and 15 (1.9%)
took the telemetered event-rate fallback — among them all solves for the
flipped point. All 800 predictions are finite; none is excluded from any
summary.

```{r fig5-dominance, fig.height=5.4}
sel.cand <- dom12b$candidate[dom12b$selected][1]
cc <- subset(contrib12b, candidate == sel.cand)
cc <- cc[order(-cc$contrib.clip.1e.15), ]
cc$rank <- seq_len(nrow(cc))
floor.y <- 1e-6
n.below <- sum(cc$contrib.clip.1e.15 < floor.y)
cc$shown <- pmax(cc$contrib.clip.1e.15, floor.y)
ggplot(cc, aes(rank, shown, color = flipped, size = flipped)) +
  geom_hline(yintercept = -log(1e-15), linetype = "dashed",
             color = col.pre) +
  geom_hline(yintercept = -log(1e-6), linetype = "dashed",
             color = col.post) +
  geom_point(alpha = 0.85) +
  annotate("text", x = 150, y = -log(1e-15) * 1.85, hjust = 0,
           label = "-log(1e-15) = 34.54 (old clip ceiling)",
           color = col.pre, size = 4) +
  annotate("text", x = 150, y = -log(1e-6) * 0.55, hjust = 0,
           label = "-log(1e-6) = 13.82 (pinned clip ceiling)",
           color = col.post, size = 4) +
  scale_y_log10(breaks = 10^c(2, 1, 0, -2, -4, -6),
                labels = c("100", "10", "1", "0.01", "1e-4", "1e-6")) +
  scale_color_manual(values = c(`FALSE` = "grey40", `TRUE` = col.flag),
                     labels = c(`FALSE` = "other 399 points",
                                `TRUE` = "deliberately flipped point"),
                     name = NULL) +
  scale_size_manual(values = c(`FALSE` = 1.6, `TRUE` = 3.4),
                    guide = "none") +
  labs(x = "points sorted by contribution (rank 1 = largest)",
       y = "per-point log-loss contribution at clip 1e-15 (log scale)") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(size = 3)))
```

**Figure 5.** Per-point log-loss contributions at the old truncation
$\varepsilon = 10^{-15}$ for the selected candidate (support 60, degree 0),
sorted in decreasing order (log y-axis; the `r n.below` points with
contributions below $10^{-6}$ are drawn at the $10^{-6}$ display floor).
The deliberately flipped point (red, rank 1) contributes $-\log(10^{-15}) =
34.54$ — $8.6\times$ the largest other contribution ($4.03$) — and is the
*only* point whose prediction lies outside $[10^{-6}, 1-10^{-6}]$, so it
accounts for 100% of the score change when the truncation moves from
$10^{-15}$ to $10^{-6}$ (a $0.052$ shift of the mean score, both
candidates). Pinning $\varepsilon = 10^{-6}$ caps that single point's
leverage at the blue ceiling; the dashed lines mark the two ceilings.

```{r table4}
t4 <- dom12b[, c("support.size", "degree", "logloss.1e.15", "logloss.1e.6",
                 "logloss.1e.3", "flip.share.of.crossclip.swing",
                 "flip.contrib.1e.15", "max.other.contrib.1e.15")]
for (cn in names(t4)[3:8]) t4[[cn]] <- sprintf("%.4f", t4[[cn]])
knitr::kable(
  t4, row.names = FALSE,
  col.names = c("support", "degree", "LL(1e-15)", "LL(1e-6)", "LL(1e-3)",
                "flip share", "flip contrib", "max other"),
  caption = tab.cap(4, paste(
    "Per-candidate log-loss scores at the three",
    "truncation levels and the single-point dominance",
    "accounting. 'flip share' is the fraction of the",
    "LL(1e-15) - LL(1e-6) score change attributable to the",
    "flipped observation (1.0 = all of it); 'flip contrib' and",
    "'max other' compare its per-point contribution at clip 1e-15",
    "with the largest contribution among the other 399",
    "points. LL(1e-6) equals the deployed selection column",
    "cv.logloss.observed exactly (residual 0).")))
```

# E2.12 study — selection stability across truncation levels

The contract treats cross-clip selection stability as a **study** — its
verdict is recorded, never gated, because near-tied candidates may
legitimately reselect when the truncation changes. The predeclared rule:
the selection is *stable* iff the log-loss minimizer at
$\varepsilon = 10^{-6}$ equals the minimizer at $\varepsilon = 10^{-3}$,
computed from the same out-of-fold predictions.

```{r fig6-crossclip, fig.height=4.6}
cs <- crossclip
cs$label <- sprintf("support %d, degree %d", cs$support.size, cs$degree)
cs$clip.f <- factor(format(cs$clip, scientific = TRUE),
                    levels = format(c(1e-15, 1e-6, 1e-3), scientific = TRUE))
mins <- do.call(rbind, lapply(split(cs, cs$clip.f), function(d) {
  d[which.min(d$cv.logloss), ]
}))
ggplot(cs, aes(clip.f, cv.logloss, color = label, group = label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  geom_point(data = mins, shape = 21, size = 6, stroke = 1.2, fill = NA,
             color = "black", show.legend = FALSE) +
  scale_color_manual(values = c(`support 60, degree 0` = col.post,
                                `support 60, degree 1` = "#8A6FB5"),
                     name = NULL) +
  labs(x = "log-loss probability truncation ε",
       y = "cross-validated log loss (lower is better)") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

**Figure 6.** Cross-validated log loss of the two E2.12(b) candidates as a
function of the truncation level $\varepsilon$ (black circles mark the
minimizer at each level). The selection is identical at $10^{-15}$ and at
the pinned $10^{-6}$ (support 60, degree 0; margin $0.0022$) and *reselects*
at $10^{-3}$ (support 60, degree 1; margin $0.0106$): the coarser truncation
caps the flipped point's penalty at $-\log(10^{-3}) = 6.9$ and also
truncates the degree-1 candidate's confident near-miss predictions, enough
to move a $0.002$-wide near-tie. The recorded study verdict is therefore
**"not stable (near-tie reselected; recorded, not gated)"**
([verdict CSV](../../../../../../../reports/e2_12_crossclip_stability_verdict.csv)) —
precisely the legitimate-reselection case the contract anticipated when it
declined to gate this comparison.

# Results summary and discussion

**E2.14 — positive.** On the near-separable fixture the pre-fix solver is
demonstrably broken in the way the plan predicted: plain Newton's deviance
increases by $\approx 489$ at step 2, oscillates, and exhausts the iteration
cap, silently degrading the prediction to the event-rate fallback. With
step-halving the same support converges in 9 iterations after a single
halving, with a monotone trajectory (largest increase $1.8\times10^{-15}$
against the $10^{-8}$ gate slack), finite coefficients, and
$\hat p = 0.125$ strictly inside $(0,1)$. Exact separation behaves exactly
as documented: bounded non-convergence at the 50-iteration cap, no `NaN`,
and a deterministic, telemetered fallback under both `unstable.action`
modes. The fix engages only where plain Newton misbehaves — on well-behaved
solves the accepted iterates are identical, and the Tier-0 E0.6 statistics
are bit-identical to the unmodified base commit.

**E2.12(a) — positive.** The train/deploy metric mismatch is not
hypothetical: on a sharp G6 surface the raw and deployed Brier scores rank
the candidates differently (margins $> 100\times$ the comparison
tolerance), and the pre-fix rule would have deployed the wrong candidate
(support 8, degree 0) over the one that is actually better as deployed
(support 60, degree 2). Post-fix, the selection column *is* the deployed
clipped metric (recomputed-equality residual $1.4\times10^{-17}$), and the
invariant is unconditional over legal bernoulli configurations: per the
independent audit's required fix (`audits/tier2_audit_2026-06-11.md`),
bernoulli now always uses the R cross-validation path — `backend = "auto"`
resolves to `"R"` exactly as for `binomial`, and an explicit C++ backend is
an error — because the C++ CV kernels return only the aggregate raw RMSE
and cannot score the deployed metric. The gate pins this resolution and a
gaussian control confirming the C++ fast path remains available outside the
binary families.

**E2.12(b) — positive, with the instability precisely localized.** At the
old $10^{-15}$ truncation, one deliberately confident-wrong held-out point
contributes $34.54$ — $8.6\times$ any other point — and owns 100% of the
score's sensitivity to the truncation level. The pin at $10^{-6}$ caps
exactly that leverage. The cross-clip study honestly records that a
$0.002$-wide near-tie reselects at $10^{-3}$: cross-clip invariance is a
robustness goal, not a correctness criterion, and this fixture sits on the
expected boundary.

**Failure modes and residual limits.** Nothing in these results validates
the gates against adversarial mutation — that is the auditor's
mutation-qualification step, deliberately not run by the implementer. All
numbers come from one machine/BLAS (macOS arm64, vecLib); the decisive
margins ($\ge 5.8\times10^{-4}$) sit far above plausible floating-point
jitter, but no second platform was exercised. The pre-fix E2.14 trajectory
is a committed replication of the pre-change update equations, not an
execution of the deleted code. Both fixtures are single deterministic
constructions (per the contract's smoke/full sizing: "deterministic — no
replication"), so no uncertainty intervals apply anywhere in this report.

# What we learned

- IRLS instability under near-separation is real but *geometrically
  selective*: it took an edge-of-cluster flipped label adjacent to a far
  same-class point to trigger it (a mid-cluster flip is benign). One
  step-halving is enough to recover a genuine fit that the pre-fix solver
  abandoned to a fallback — the fix improves fits, not just telemetry.
- The variance floor and $\pm 35$ linear-predictor clamp already prevented
  `NaN`; what they could not prevent was non-monotone oscillation, which is
  invisible at the endpoint and only assertable on the trajectory. Gating
  the *trajectory* was the right call.
- Metric mismatches between selection and deployment can flip model choice
  on realistic sharp-boundary binary surfaces — 16% of out-of-fold
  predictions leaving $[0,1]$ was enough. Selection now optimizes what is
  deployed.
- A $10^{-15}$ log-loss truncation converts a single all-zero-neighborhood
  fallback prediction into the dominant term of the selection score; a
  $10^{-6}$ pin bounds that single-point leverage while leaving every
  in-range contribution untouched.
- A near-tie ($\Delta \approx 0.002$) between candidates can legitimately
  reselect under truncation changes; recording this as a study verdict
  rather than gating it keeps the acceptance criteria honest.

# Appendix — reproducibility

All commands assume the package root of worktree
`/Users/pgajer/current_projects/geosmooth-t2`, branch
`codex/geosmooth-t2-binary-hygiene`.

**Report build:** `r build.time`. **Result generation:**
`r mf("results_generated_eastern")` (UTC `r mf("results_generated_utc")`) at
commit `` `r mf("git_head")` ``; BLAS `` `r basename(mf("blas"))` ``;
`r mf("r_version")`. Bundle cross-check at generation:
`r mf("bundle_crosscheck")`.

Result artifacts consumed (committed; report-specific inputs under
`inputs/` beside this report with SHA-256 in the
[input manifest](inputs/tier2_report_inputs_manifest.txt); gate/study
evidence at the package-level `reports/`):

- E2.14 inputs: [`e2_14_fixture.csv`](inputs/e2_14_fixture.csv),
  [`e2_14_deviance_traces.csv`](inputs/e2_14_deviance_traces.csv),\
  [`e2_14_solver_summary.csv`](inputs/e2_14_solver_summary.csv),
  [`e2_14_fallback_telemetry.csv`](inputs/e2_14_fallback_telemetry.csv);
- E2.12 inputs: [`e2_12_fixture_a.csv`](inputs/e2_12_fixture_a.csv),
  [`e2_12a_selection_metric.csv`](inputs/e2_12a_selection_metric.csv),\
  [`e2_12b_contributions.csv`](inputs/e2_12b_contributions.csv),
  [`e2_12b_clip_dominance.csv`](inputs/e2_12b_clip_dominance.csv),\
  [`e2_12b_fit_accounting.csv`](inputs/e2_12b_fit_accounting.csv);
- gate/study evidence (`reports/`):
  [`e2_14_prefix_newton_overshoot.csv`](../../../../../../../reports/e2_14_prefix_newton_overshoot.csv),\
  [`e2_12_crossclip_scores.csv`](../../../../../../../reports/e2_12_crossclip_scores.csv),
  [`e2_12_crossclip_stability_verdict.csv`](../../../../../../../reports/e2_12_crossclip_stability_verdict.csv).

Generation and rendering commands:

```sh
R="dev/methods/lps/reports/tier2/binary_hygiene/2026-06-11"

# 1. (re)generate every report-input CSV + manifest into $R/inputs/
#    (deterministic; the optional argument cross-checks against an
#    execution bundle):
Rscript "$R/scripts/export_tier2_report_inputs.R" \
  audit_artifacts/tier2_20260611T084642Z

# 2. pre-fix trajectory + cross-clip study artifacts (committed
#    gate/study evidence under reports/):
Rscript validation/e2_14_prefix_newton_overshoot.R
Rscript validation/e2_12_crossclip_stability_study.R

# 3. render this report, HTML + PDF (written into $R/):
Rscript "$R/scripts/render_lps_tier2_binary_hygiene_report.R"

# acceptance evidence proper (clean committed tree; auditor-facing):
EXECUTOR="<id>" bash scripts/ci/run_tier2_execution_artifact.sh
```

Seeds and determinism: the E2.14 fixtures are seed-free closed-form
constructions; E2.12 uses `set.seed(2121)` (sub-item a) and
`set.seed(7001)` with the label of observation 7 flipped (sub-item b);
cross-validation folds are the explicit `foldid = rep(1:5, length.out =
400)` in all fits (never the internal `cv.seed`).

Related committed documents and machine-local evidence:

- gate test files:
  [`test-lps-binary-separation.R`](../../../../../../../tests/testthat/test-lps-binary-separation.R),\
  [`test-lps-binary-metric-consistency.R`](../../../../../../../tests/testthat/test-lps-binary-metric-consistency.R);
- implementer handoffs:
  [`e2_14_implementer_handoff_2026-06-11.md`](../../../../../../../phase_handoffs/e2_14_implementer_handoff_2026-06-11.md),\
  [`e2_12_implementer_handoff_2026-06-11.md`](../../../../../../../phase_handoffs/e2_12_implementer_handoff_2026-06-11.md);
- execution bundles (machine-local, gitignored):
  `audit_artifacts/tier2_20260611T075822Z` and\
  `audit_artifacts/tier2_20260611T084642Z`.
