\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{xurl}
\usepackage{enumitem}
\usepackage{microtype}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!50!black,
  urlcolor=blue!55!black,
  citecolor=blue!50!black
}

\input{build_info.tex}

\newcommand{\R}{\mathbb{R}}
\newcommand{\GCV}{\operatorname{GCV}}
\newcommand{\RMSE}{\operatorname{RMSE}}

\title{PS-LPS Progress Report}
\author{%
  \small
  \begin{tabular}{@{}c@{}}
  Source path: \texttt{\string~/current\_projects/geosmooth/split\_handoffs/}\\
  \texttt{ps\_lps\_progress\_2026-06-05/ps\_lps\_progress\_report.tex}
  \end{tabular}%
}
\date{Build: \reportbuilddatetime}

\begin{document}
\maketitle

\section*{Purpose}

This report records the current state of prediction-synchronized local
polynomial smoothing, abbreviated \emph{PS-LPS}.  It is intended as a living
progress report rather than a final method paper.  The design document answers
the question ``what is PS-LPS?''  This report answers:
\begin{quote}
What has been implemented, what has been observed, what remains uncertain, and
what should be audited or tested next?
\end{quote}

The report is deliberately conservative.  The first-batch results are
encouraging, but they are preliminary until the implementation and experiment
have passed an independent audit and have been tested on a broader validation
suite.

\section{Motivation}

Ordinary local polynomial smoothing (LPS) fits one local polynomial model around
each anchor point.  This makes LPS simple and often effective, but it also means
that neighboring local charts are fitted independently even though their
supports overlap heavily.  Two charts can make different predictions at the same
overlap point, and ordinary LPS has no explicit mechanism to reduce that
incoherence.

Before PS-LPS, we tested whether the sum of ordinary independent-chart local
GCV scores could select global LPS parameters.  That first experiment did not
support using
\[
  \sum_i \GCV_i
\]
as a standalone global selector for independent LPS.  The diagnostic often
favored high local degrees-of-freedom ratios and did not reliably track Truth
RMSE on the first-batch examples.

PS-LPS takes a different route.  Instead of using local GCV alone to select
parameters, it modifies the smoother itself by adding a prediction
synchronization term.  The central idea is that overlapping local charts should
not merely fit their own local responses; they should also agree where their
prediction domains overlap.

\section{Model Definition}

Let \(x_1,\ldots,x_n\in\R^D\) be observed covariate points and let \(y_i\) be
the observed response at point \(x_i\).  For each anchor \(i\), define a local
support
\[
  N_i\subset\{1,\ldots,n\},
  \qquad |N_i|=k.
\]
Let \(w_{ij}\ge 0\) be the kernel weight assigned by chart \(i\) to point
\(j\in N_i\).  Let
\[
  \phi_i(x_j)\in\R^q
\]
be the local polynomial feature vector for point \(x_j\), expressed in the
local chart centered at anchor \(i\).  The local coefficient vector at anchor
\(i\) is \(\beta_i\in\R^q\).

The local data-fit loss is
\[
  \mathcal L_{\mathrm{local}}(\beta)
  =
  \frac12
  \sum_{i=1}^{n}
  \sum_{j\in N_i}
  w_{ij}
  \{y_j-\phi_i(x_j)^\top\beta_i\}^2.
\]

For two charts \(i\) and \(\ell\), define the overlap
\[
  O_{i\ell}=N_i\cap N_\ell.
\]
The PS-LPS synchronization penalty is
\[
  \mathcal S(\beta)
  =
  \frac12
  \sum_{(i,\ell)}
  \sum_{r\in O_{i\ell}}
  \omega_{i\ell r}
  \left\{
    \phi_i(x_r)^\top\beta_i
    -
    \phi_\ell(x_r)^\top\beta_\ell
  \right\}^2.
\]

The PS-LPS estimate is
\[
  \widehat\beta
  =
  \arg\min_{\beta_1,\ldots,\beta_n}
  \left[
    \mathcal L_{\mathrm{local}}(\beta)
    +
    \lambda_{\mathrm{sync}}\mathcal S(\beta)
  \right].
\]
When \(\lambda_{\mathrm{sync}}=0\), the model should reduce to independent LPS
local chart fits with the same support, kernel, degree, and chart dimensions.
This is an important audit gate for the implementation.

\section{Overlap Weight Decision}

The current recommended overlap weight is the normalized product of the two
ordinary chart weights:
\[
  \omega_{i\ell r}
  =
  |O_{i\ell}|
  \frac{
    w_{ir}w_{\ell r}
  }{
    \sum_{s\in O_{i\ell}} w_{is}w_{\ell s}+\epsilon
  }.
\]
The product \(w_{ir}w_{\ell r}\) emphasizes points that are central in both
charts.  This matches the intended geometry: charts should be synchronized most
strongly in the middle of their overlap, and less strongly near either chart's
boundary.  The pair normalization makes the overall synchronization scale more
stable across chart pairs, kernels, and overlap sizes.

\section{Implementation Status}

The first implementation is an R reference prototype in
\[
  \texttt{\string~/current\_projects/geosmooth/R/ps\_lps.R}.
\]
The exported user-facing function is
\[
  \texttt{fit.ps.lps()}.
\]

The prototype currently:
\begin{itemize}[leftmargin=*]
  \item uses ordinary local PCA charts;
  \item supports fixed scalar chart dimension or a local chart-dimension vector;
  \item fits a fixed support size, kernel, degree, and chart-dimension policy;
  \item tunes only \(\lambda_{\mathrm{sync}}\) by CV in the first experiment;
  \item assembles a sparse global least-squares system;
  \item solves sparse normal equations with a small scale-relative numerical
    ridge.
\end{itemize}

The current numerical ridge is a prototype stabilization device.  It should be
audited and eventually reported as a solver diagnostic, especially when testing
whether \(\lambda_{\mathrm{sync}}=0\) reproduces ordinary LPS.

\section{Diagnostics}

The PS-LPS implementation and first experiment record the following quantities:
\begin{itemize}[leftmargin=*]
  \item CV RMSE, computed on held-out responses;
  \item Observed RMSE, computed against the noisy observed response;
  \item Truth RMSE, computed against synthetic truth when available;
  \item synchronization energy \(\mathcal S(\widehat\beta)\);
  \item mean synchronization disagreement;
  \item local degrees-of-freedom ratios;
  \item total synchronized local GCV.
\end{itemize}

The synchronized local GCV diagnostic is defined after solving the synchronized
problem.  For chart \(i\),
\[
  \operatorname{RSS}^{\mathrm{PS}}_i
  =
  \sum_{j\in N_i}
  w_{ij}
  \{y_j-\phi_i(x_j)^\top\widehat\beta_i\}^2.
\]
Let \(r_i\) be the numerical rank of the local weighted design matrix.  Then
\[
  \GCV^{\mathrm{PS}}_i
  =
  \frac{\operatorname{RSS}^{\mathrm{PS}}_i/k}
       {(1-r_i/k)^2},
  \qquad
  \GCV^{\mathrm{PS}}_{\Sigma}
  =
  \sum_i \GCV^{\mathrm{PS}}_i.
\]
This is not the same as total local GCV for independent LPS charts.  The
residuals are computed from the synchronized chart coefficients
\(\widehat\beta_i\).  This distinction matters because the preliminary results
suggest that \(\GCV^{\mathrm{PS}}_{\Sigma}\) may be more aligned with Truth RMSE
than the independent-LPS version.

\section{First-Batch Experiment}

The first PS-LPS experiment used the frozen first-batch non-manifold datasets:
\[
  \texttt{\string~/current\_projects/geosmooth/split\_handoffs/}
\]
\[
  \texttt{lps\_local\_auto\_nonmanifold\_first\_batch\_2026-06-05/}.
\]
The generated PS-LPS report is:
\[
  \texttt{\string~/current\_projects/geosmooth/split\_handoffs/}
\]
\[
  \texttt{ps\_lps\_first\_batch\_experiment\_2026-06-05/}
\]
\[
  \texttt{ps\_lps\_first\_batch\_experiment\_report.html}.
\]

The comparison included four methods:
\begin{itemize}[leftmargin=*]
  \item ordinary LPS with \texttt{chart.dim = "auto"};
  \item ordinary LPS with \texttt{chart.dim = "local.auto"};
  \item PS-LPS using the ordinary \texttt{auto} source settings;
  \item PS-LPS using the ordinary \texttt{local.auto} source settings.
\end{itemize}
In the first experiment, PS-LPS reused the corresponding ordinary LPS selected
support size, kernel, degree, and chart-dimension policy, and tuned only
\[
  \lambda_{\mathrm{sync}}\in\{0,0.1,1\}.
\]

\section{Audit Result and Response}

The first implementation audit was blocked for two substantive reasons.  First,
the implementation used an unconditional ridge in the normal-equation solve, so
\(\lambda_{\mathrm{sync}}=0\) did not reproduce ordinary LPS fixed-chart fits.
Second, the synchronization diagnostics were incorrectly reported as zero when
\(\lambda_{\mathrm{sync}}=0\), even though the unmultiplied disagreement
\(\mathcal S(\widehat\beta)\) is generally nonzero.

The response was to make the ridge explicit rather than hidden.  The current
implementation has a user-facing parameter
\[
  \lambda_{\mathrm{ridge}}\ge 0.
\]
When \(\lambda_{\mathrm{ridge}}=0\) and
\(\lambda_{\mathrm{sync}}=0\), PS-LPS is tested to reproduce ordinary
independent LPS full-data fixed-chart fitted values.  When
\(\lambda_{\mathrm{ridge}}>0\) and \(\lambda_{\mathrm{sync}}=0\), the method is
instead an independent ridge-stabilized LPS fit.  Therefore any
ridge-stabilized PS-LPS comparison must include the matched ridge-LPS baseline.

The diagnostics now compute and report the unmultiplied synchronization energy
\[
  \mathcal S(\widehat\beta)
\]
for every fitted coefficient vector whenever overlap rows exist, including the
\(\lambda_{\mathrm{sync}}=0\) case.  The ambiguous diagnostic name
\texttt{mean.sync.disagreement} has been retained only as a compatibility alias;
the preferred diagnostic is
\[
  \texttt{mean.sync.squared.disagreement}.
\]

\section{Focused Tests}

The focused PS-LPS test file is
\[
  \texttt{\string~/current\_projects/geosmooth/tests/testthat/test-ps-lps.R}.
\]
It currently checks:
\begin{itemize}[leftmargin=*]
  \item normalized product overlap weights have the intended per-pair mass;
  \item \(\lambda_{\mathrm{ridge}}=0\) and
    \(\lambda_{\mathrm{sync}}=0\) reproduce ordinary LPS fixed-chart fitted
    values;
  \item positive ridge with \(\lambda_{\mathrm{sync}}=0\) reproduces the
    independent ridge-LPS solver path;
  \item synchronization energy and mean squared disagreement are nonzero
    diagnostics at \(\lambda_{\mathrm{sync}}=0\) when charts disagree.
\end{itemize}
The targeted test run passed with 9 expectations and no failures.

\section{Refined First-Batch Experiment}

The refined report is:
\[
  \texttt{\string~/current\_projects/geosmooth/split\_handoffs/}
\]
\[
  \texttt{ps\_lps\_first\_batch\_refined\_experiment\_2026-06-05/}
\]
\[
  \texttt{ps\_lps\_first\_batch\_refined\_experiment\_report.html}.
\]

The refined comparison separates six methods:
\begin{itemize}[leftmargin=*]
  \item ordinary LPS \texttt{auto};
  \item ridge-LPS \texttt{auto};
  \item PS-LPS \texttt{auto};
  \item ordinary LPS \texttt{local.auto};
  \item ridge-LPS \texttt{local.auto};
  \item PS-LPS \texttt{local.auto}.
\end{itemize}
All ridge-LPS and PS-LPS rows use
\[
  \lambda_{\mathrm{ridge}}=10^{-8}.
\]
The PS-LPS grid remains
\[
  \lambda_{\mathrm{sync}}\in\{0,0.1,1\},
\]
and ridge-LPS is the matched \(\lambda_{\mathrm{sync}}=0\) baseline.

All 84 method--dataset rows completed successfully.  The best Truth-RMSE method
counts over the 14 first-batch datasets were:
\begin{center}
\begin{tabular}{@{}lr@{}}
\toprule
Method & Number of datasets where best \\
\midrule
PS-LPS local.auto & 8 \\
PS-LPS auto & 5 \\
ridge-LPS local.auto & 1 \\
ordinary LPS variants & 0 \\
ridge-LPS auto & 0 \\
\bottomrule
\end{tabular}
\end{center}

Median Truth RMSE by method was:
\begin{center}
\begin{tabular}{@{}lr@{}}
\toprule
Method & Median Truth RMSE \\
\midrule
PS-LPS local.auto & \(0.04048\) \\
PS-LPS auto & \(0.04517\) \\
ridge-LPS local.auto & \(0.04868\) \\
LPS local.auto & \(0.05755\) \\
ridge-LPS auto & \(0.05752\) \\
LPS auto & \(0.06978\) \\
\bottomrule
\end{tabular}
\end{center}

The refined delta decomposition is the main interpretive result.  The median
Truth-RMSE changes were:
\begin{center}
\begin{tabular}{@{}llr@{}}
\toprule
Method & Reference & Median Truth-RMSE delta \\
\midrule
ridge-LPS auto & ordinary LPS auto & \(-0.00651\) \\
ridge-LPS local.auto & ordinary LPS local.auto & \(-0.00638\) \\
PS-LPS auto & ordinary LPS auto & \(-0.02437\) \\
PS-LPS local.auto & ordinary LPS local.auto & \(-0.02190\) \\
PS-LPS auto & ridge-LPS auto & \(-0.01048\) \\
PS-LPS local.auto & ridge-LPS local.auto & \(-0.00865\) \\
\bottomrule
\end{tabular}
\end{center}
Thus the initial favorable result was partly a ridge-stabilization effect, but
not only a ridge effect.  In this refined first-batch experiment, positive
prediction synchronization still improved over the matched ridge-LPS baseline
in median Truth RMSE.  The dataset-wise caveat is that each PS-LPS variant
improved over its matched ridge-LPS baseline on 13 of 14 datasets, not on every
dataset.  Therefore the appropriate short summary is ``median improvement over
matched ridge-LPS, with 13/14 dataset-wise wins for each PS variant.''

\section{Open Questions}

The following questions remain open and should guide the next phase:
\begin{itemize}[leftmargin=*]
  \item Is the improvement over ridge-LPS stable under a wider
    \(\lambda_{\mathrm{sync}}\) grid?
  \item How sensitive are the conclusions to the ridge scale
    \(\lambda_{\mathrm{ridge}}\)?
  \item Does PS-LPS still improve on prospective datasets that were not used to
    motivate the method?
  \item Does total synchronized local GCV remain aligned with CV RMSE and Truth
    RMSE on larger suites?
  \item Should PS-LPS select support size, kernel, degree, and chart dimension
    directly, or remain a synchronization refinement around an ordinary
    LPS-selected candidate?
  \item Is the R prototype sufficient for near-term experiments, or should the
    sparse system assembly and solve be moved to C++?
\end{itemize}

\section{PS-LPS-S1 Sensitivity Plan}

The re-audit accepted the refined ridge-stabilized contract with minor
follow-ups.  The next validation phase is therefore not yet a prospective
experiment.  The immediate phase is:
\[
  \text{PS-LPS-S1: first-batch lambda/ridge sensitivity}.
\]
Its purpose is to test whether the refined first-batch PS-LPS signal is stable
to the two prototype tuning axes:
\[
  \lambda_{\mathrm{sync}}
  \quad\text{and}\quad
  \lambda_{\mathrm{ridge}}.
\]

The planned synchronization grid is
\[
  \lambda_{\mathrm{sync}}
  \in
  \{0,\ 0.01,\ 0.03,\ 0.1,\ 0.3,\ 1,\ 3,\ 10\},
\]
and the planned ridge grid is
\[
  \lambda_{\mathrm{ridge}}
  \in
  \{0,\ 10^{-10},\ 10^{-8},\ 10^{-6}\}.
\]
For every ridge scale, the matched baseline is the candidate with
\(\lambda_{\mathrm{sync}}=0\) and the same support size, kernel, degree, chart
rule, and ridge scale.  When \(\lambda_{\mathrm{ridge}}=0\), this is ordinary
unregularized LPS.  When \(\lambda_{\mathrm{ridge}}>0\), this is ridge-LPS.

The S1 report should include:
\begin{itemize}[leftmargin=*]
  \item selected \(\lambda_{\mathrm{sync}}\) by dataset, chart rule, and ridge
    scale;
  \item candidate-level Truth RMSE over the
    \((\lambda_{\mathrm{ridge}},\lambda_{\mathrm{sync}})\) grid;
  \item selected Truth-RMSE deltas against the matched
    \(\lambda_{\mathrm{sync}}=0\) baseline;
  \item median and dataset-wise win summaries;
  \item total local GCV, synchronization energy, and mean squared overlap
    disagreement diagnostics.
\end{itemize}
The decision criterion is conservative: proceed to prospective PS-LPS
experiments only if the improvement over matched ridge-LPS is not confined to a
single ridge scale or to a boundary value of \(\lambda_{\mathrm{sync}}\).

\section{PS-LPS-S1 Results}

The S1 experiment has now been run on the frozen first-batch non-manifold
suite.  The report is:
\[
\begin{gathered}
  \texttt{~/current\_projects/geosmooth/split\_handoffs/}\\
  \texttt{ps\_lps\_s1\_lambda\_ridge\_sensitivity\_2026-06-05/}\\
  \texttt{ps\_lps\_s1\_lambda\_ridge\_sensitivity\_report.html}.
\end{gathered}
\]
The run evaluated \(896\) candidate rows:
\[
  14\ \text{datasets}
  \times 2\ \text{chart rules}
  \times 4\ \lambda_{\mathrm{ridge}}\text{ values}
  \times 8\ \lambda_{\mathrm{sync}}\text{ values}.
\]
All candidate rows completed with status \texttt{ok}.  The requested
\(\lambda_{\mathrm{ridge}}=0\) rows are included, but the numerical solver did
use its internal fallback ridge on a small number of zero-ridge candidate
solves.  The S1 tables record the realized \(\lambda_{\mathrm{ridge}}\)
diagnostics as \texttt{ridge\_median} and \texttt{ridge\_max}.  Thus the
zero-ridge rows are valuable as a stress test of the no-ridge path, but they
should not be described as uniformly exact unregularized normal-equation solves.

The selected-by-CV median Truth-RMSE deltas against the matched
\(\lambda_{\mathrm{sync}}=0\) baseline were:
\begin{center}
\begin{tabular}{@{}lrrr@{}}
\toprule
Chart rule & \(\lambda_{\mathrm{ridge}}\) & Median delta & Wins \\
\midrule
\texttt{auto} & \(0\) & \(-0.02224\) & \(12/14\) \\
\texttt{local.auto} & \(0\) & \(-0.01523\) & \(13/14\) \\
\texttt{auto} & \(10^{-10}\) & \(-0.01319\) & \(12/14\) \\
\texttt{local.auto} & \(10^{-10}\) & \(-0.01104\) & \(12/14\) \\
\texttt{auto} & \(10^{-8}\) & \(-0.01523\) & \(12/14\) \\
\texttt{local.auto} & \(10^{-8}\) & \(-0.01095\) & \(12/14\) \\
\texttt{auto} & \(10^{-6}\) & \(-0.01425\) & \(12/14\) \\
\texttt{local.auto} & \(10^{-6}\) & \(-0.01057\) & \(12/14\) \\
\bottomrule
\end{tabular}
\end{center}
Negative deltas mean that the selected synchronized fit has lower Truth RMSE
than the matched independent baseline at the same ridge scale.  The signal is
therefore not confined to one ridge scale: every ridge scale and both chart
rules show a negative median delta.

The caution is that the selected synchronization strength is usually at the
upper boundary of the S1 grid.  Across all ridge scales, the median selected
\(\lambda_{\mathrm{sync}}\) was \(10\) for both chart rules.  This does not
invalidate the positive S1 result, but it does mean that the next sensitivity
run should extend the synchronization grid above \(10\) and should examine
whether the Truth-RMSE profile plateaus, improves further, or eventually
deteriorates.  In short, S1 supports prediction synchronization as a useful
refinement, but it does not yet freeze the synchronization search range.

\section*{Summary}

PS-LPS has moved from a concept note to a tested ridge-stabilized prototype.
The blocked audit identified a real interpretability problem, and the refined
experiment now separates ordinary LPS, ridge-LPS, and synchronized PS-LPS.  The
first-batch evidence remains encouraging because PS-LPS improves over the
matched independent baseline in median Truth RMSE across the tested ridge
scales.  The main limitation is now different: S1 frequently selects the largest
tested synchronization weight, so the next experiment should extend
\(\lambda_{\mathrm{sync}}\) before prospective validation.

\end{document}
